Learning and cognition in a decision made at reflex speed

Version 1

Abstract

Abstract

In recent years it has become clear that many decisions do not obey the rule that more time yields better decisions. These decisions can be made remarkably fast and yet accurately, sometimes based on very limited information. It is presently unclear whether such ‘blink’ or high-speed decisions lack cognitive aspects that only much slower ‘deliberative’ decision-making can support. Here we demonstrate an unexpected degree of flexibility and cognition in a decision made by a hunting animal at reflex-like speed. Based on observing initial speed, direction, and height of falling prey archerfish decide in just 40 milliseconds on a turn toward the later ballistic landing point. This enables the fish to dash off to arrive simultaneously with prey and to secure it against numerous competitors. We established an approach that allowed us to replace ballistics, the rule that governs the turn decisions, with a novel rule of how to connect the input variables with the rewarded turns. This approach revealed that the fish are not using a hardwired circuit but were able to reprogram their decision in efficient ways that allowed them to immediately generalize to untrained settings. Training even allowed the decision to simultaneously use two distinct sets of rules, one for each of two distinct objects. The flexibility of the decision and the occurrence of high-level cognitive features are counterintuitive for a reflex-like decision made faster than an Olympic sprinter can respond to the start gun. However, they imply that combining speed and accuracy in rapid decisions does not generally make them less smart than decisions made over far longer timescales.

Introduction

Everyday experience suggests that the quality of our decisions depends on how much time and effort went into them. However, many decisions drastically defy this rule (e.g. 1-7). Although diverse, they have in common that they all can be made surprisingly fast, often based on limited information and yet can be accurate, often with no increase in accuracy when allowed more time. For instance, rats categorize odors reliably within one sniff, i.e. within about 300 ms. Allowing for longer sampling time did not increase accuracy even for difficult tasks (3,8,9). Similar short sampling has been described for visual perceptual decisions. Humans can reliably classify natural images after surprisingly brief viewing times of only 20 to 40 ms (1,10) and it has long been known, though for comparatively simple tasks, that extending viewing time did not affect the thresholds in visual decisions that are based on stereoscopic depth perception (11,12), velocity-discrimination (13,14), or contrast sensitivity (15,16). This suggests that a natural sampling interval – the time it takes to produce a sniff, or a saccade – is the elementary interval in many visual and olfactory perceptual decisions (3). Adding urgency into laboratory decision tasks by having the subject respond to sensory information after it has already launched movement (e.g. a saccade) has allowed to dissect perceptual and motor aspects of decision-making and thereby allowed the actual speed of perceptual decisions to be measured with high temporal accuracy (5,17). Outside the laboratories, the use of rapid decisions is probably very common in sports. For instance, baseball hitters have a mere 400 ms to move their bat to an ideal hit point based on an estimated 70 ms of visual processing to inform the movement of the bat (18–20), with a mere 40 ms of more or lesser sampling deciding between the best achievable and a bad motor choice (5). For some much slower and more complex human ‘blink’ decisions, such as which product to buy, it has been shown that blink decisions can be superior compared to those made much slower and based on more information (6,7,21–23).

An intriguing question about these diverse but impressively fast and accurate decisions is whether they lack aspects that only slower decisions can offer. For instance, high-speed decisions might be fast because they make clever use of simplifying heuristics and of hardwired neuronal machinery so that they work only under the specific conditions for which they were optimized. Are slower ‘deliberate’ decisions therefore needed when flexibility and perhaps higher cognitive aspects are required? While this is presently poorly understood, many behavioral findings – from the movement of baseball outfielders and frisbee-catching dogs (24, 25) to the chase of hoverflies (26) – are available that could be (and have been 6,7) interpreted as supporting such a conclusion. For instance, to intercept a passing female, male hoverflies do not continuously monitor the female to guide their own movement but rapidly decide on an interception course immediately when a passing female comes into sighting distance (26). Based on information taken at this time the male immediately selects a specific turn so that moving in the chosen direction at fixed acceleration leads straight to the point of catch. This decision is based on hardwired assumptions that appear to be genetically imprinted in the neural circuits that govern the initial turn (Fig. 1A). When any of these assumptions were experimentally violated, for example when the target had unusual speed or sighting distance, then the turns were massively in error just as was predicted (26). So, the hoverflies solution affords a striking degree of speed and accuracy but clearly comes at the cost of flexibility. Nevertheless, there appear to be no selective pressures for the decision to afford more flexibility, in fact, the apparent limitations are potentially even to the advantage of the hoverfly males (26).

High-speed decisions used by animals to catch objects.

(A) Male hoverflies use a rapid open loop-strategy to catch passing females: When sighting a female moving at speed ‘v’ at fixed distance ‘D’ the male rapidly turns in a direction in which it will, always accelerating at a fixed rate ‘a’, be led to a point at which it can intercept the female. Here, information is taken only once. To make this decision, the male uses hardwired circuitry that assumes fixed values for D, v and a, which allows to simplify the complex problem considerably (6). (B) To secure downed prey hunting archerfish also use an open-loop strategy: Based only on visual information sampled after prey starts moving the fish turn to the later impact point and then (corresponding points in time indicated by an asterisk) take off at a speed that makes them arrive at full speed very slightly after their prey. The turns are produced in a so-called C-start maneuver and can be initiated in as little as 40 ms. The hunting context does not allow similar simplifications as in the hoverfly and fish can respond from a large range of distances and for any orientation. They are also, for instance, not limited to targets of an expected size or speed. (C) When prey is made to slide purely horizontally on a glass plate – without actually falling downwards – then the fish still elicit turn decisions. These are not aimed at the fly’s actual position and also not in the direction of the fly’s movement, but at the ‘virtual’ impact point (VIP), where ballistically falling prey would have landed given its initial speed, direction and height.

Here we addressed the question of limits in flexibility in a high-speed decision that is made faster than an Olympic sprinter responds to the start gun. Like the decisions of hoverflies (26), batters (18–20) or outfielders (24,25) it is also a decision based on visual information and it is also about being at the right spot at the right time. It operates under interesting constraints and has properties that make it amenable to experimental dissection in the lab (27). The decision is an essential part of the hunting behavior of archerfish, that are renowned for shooting down aerial prey with jets of water (28–30) (Fig. 1B): As soon as they have dislodged a prey insect with their shot, the fish (shooter and bystander archerfish) have the little time that remains till prey hits the water surface to be at the point of catch. A failure to arrive at this spot simultaneously with their prey means losing it to one of their numerous competitors, large crowds of other surface-feeding fish that also can launch pre-impact starts based on visually observing how the target falls (31). To outperform these competitors, archerfish watch the initial movement of a falling prey – which is typically hit from the side (see Fig. 1B) – and then select a turn right to the later landing point (e.g. 27, 32): Additionally, they then dart off with the speed needed to arrive just in time at the right spot (33,34). Because aerial prey is scarce (31), archerfish engage prey over a wide range of distances (30,31) and, hence, must also be able to select the correct turns for a wide range of sighting distances of prey, vertical and horizontal initial speed levels of prey, prey sizes. Archerfish can therefore not use similar simplifications that work so well for hoverflies (26). Indeed, the fish can decide correctly even when these variables were varied independently from each other over large ranges (27,35,36). Of key importance for our present study is this interesting property of the decision (Fig. 1C): Suppose a prey object is made, by a stream of air from a blastpipe (32), to slide on top of a glass plate. Faced with such a condition, hunting archerfish do not follow the trajectory of the object. Instead, they still make their turn decisions just as they would in absence of the glass plate and accurately turn towards the ballistic landing point predicted from the initial height, initial direction and initial speed of the sliding target (32). It should be stressed and will become clear in the results, that the accuracy of the turn decisions allows to decide whether the turns were to the ballistic landing point, or another point, such as the last sighting of the fly before turn onset (32) or the landing point appropriate for another estimate of initial height (35, 36). The glass plate experiments therefore show that the turn decision maps the initial values of motion of prey to the turn the fish must make and that this mapping is appropriate for ballistically falling objects. In these experiments prey movement was independent of shooting, and it turned out later, that the decision does not use prior information – i.e. on target position, height, timing of motion onset – that would be available already before prey starts its trajectory. This was tested in experiments in which size, initial position and height were changed relative to what the fish could have expected (27,35,36). Furthermore, accuracy and latency of the turn decisions are independent from viewing distance and the orientation of target motion relative to the responding fish (35,36).

Although exhibiting impressive features, the ‘machinery’ behind the turn decisions could still be hardwired and somehow restricted to work only for ballistically falling prey, such that any combination of input variables (speed, direction, height) and coordinates of the fish leads to the turn that would be appropriate for an object that falls ballistically. Such a hardwired mechanism could be acquired over the course of evolution or could be acquired or refined during early ontogeny. But once established it would be fixed. This would explain why all archerfish studied in our lab during the last 20 years consistently showed robust turn decisions to ballistic landing points (27,31–37). It would also be according to what intuition suggests: reflex-like speed should exclude any further flexibility. Finally, smaller prey is completely wetted by the shots and falls ballistically (32,35), so that no further flexibility would seem to be required, ecologically, from the decision. However, as we report here, adult fish readily learn to extract a new artificially introduced rule of how to connect the three independent input variables to the rewarded turns and to drive their high-speed decisions according to this new rule, ignoring all their ballistics. Moreover, they re-programmed their decisions in surprisingly clever ways that not only allowed them to immediately generalize their high-speed decisions to untrained constellations but even to simultaneously handle two different rules.

Results

A technique to substitute ballistics with a new rule

If the archerfish’s high-speed decision was hardwired it would fail after a sudden and drastic change in how the rewarded output needs to be linked to the input. Adult archerfish would then continue to turn to the appropriate ballistic impact point and would have to learn ways to correct their error later on their approach path. Testing this idea requires a way to exchange the ballistic rule with a new rule and to reward the fish at locations that are dictated by this new rule. The responses to purely planar movement (33) (Fig. 1C) and the independence of the decision of the shooting context (28,33,36) suggests that LCD monitors could, in principle, be used to provide all information archerfish use for their decision. To reward the fish at defined virtual impact points without additional confounding motion cues we developed automatic feeders that can deliver food according to experimenter-defined rules of how to link place and time of food delivery to initial target motion (Fig. 2A, Fig. S2). However, these feeders are prominent landmarks and as such could substantially affect the decision. Surprisingly, a series of experiments showed, that the feeders did not influence the decision at all and that it was still exclusively based on initial target motion (Fig. 2B). Operating the feeders without showing motion on the screen failed to elicit decisions. Most importantly, when we created a direct conflict between aiming at a feeder or aiming at the ballistic virtual impact point (VIP) – appropriate for the fish’s position and the target’s movement – then the fish aimed at the VIPs, not at the feeders even though this was where they got their reward (Fig. 2B; 15° offset, n=353; 30° offset, n=198). Moreover, because these tests were conducted so that the VIPs also differed relative to other potential landmarks in the tank, the fish were also not starting to positions marked by landmarks other than the feeders. In other words, even in presence of the feeders the decisions were only based on the movement information on the screen and were not influenced by cues from the feeders or from any other landmarks in the tank.

Establishing a technique that allows to replace ballistics.

To define arbitrary relations between input and the rewarded turns, as required in this study, we first established and calibrated a virtual reality approach. (A) Sketch of setup with motion presented on a screen and feeders operated to deliver food in time at the VIPs. Evidence that the decisions are not influenced by the feeders: Same errors to VIPs with feeders present or absent. Inset illustrates how error of turn decision is defined (for details see Fig. S1). No rapid turns are elicited when feeders are operated without movement on screen. When motion directions are chosen to create a conflict between turning either to the VIPs or to the feeder positions (offset either 11.2 or 22.3 cm), then turns minimized the error to the VIPs but not to the feeder positions. Data are represented as medians. (C-G) Comparing the nature of the turn decisions with real and virtual impact points. Same errors (C) to virtual and real impact point, same latency (D), same kinematics (E, duration of bending phase). (F, G) Under virtual conditions accurate starts are also possible across wide ranges of distance (F) and turning angle (G). See Figs. S1, S2, Movies S1, S2 and text for detailed statistics. ***p<0.001, n.s. not significant.

Using this system, it should be possible to achieve our aim of presenting fish with novel non-ballistic rules that connect the input variables and the rewarded locations. However, one problem remains that needs to be examined: Given the short time in which the fish must respond and given the comparably low frame rate of the screen, it is unclear whether the decisions made with target movement displayed on the LCD screens would be comparable to those made with real prey. For instance, they could be much less accurate, be slower, or only possible from specific positions. To address this, we directly compared the performance of the same fish in response to either real falling objects (see Movie S1, that illustrates the speed of the response) or to the VR stimuli (Movie S2). Given the limitations in the temporal resolution of the screen it is striking that no aspect of the decisions was statistically different under real and virtual conditions (Fig. 2C-G, n=101 (real) and n=85 (virtual) starts). The accuracy of the aim achieved at the end of the turns was not statistically different (p=0.506, Mann-Whitney; Fig. 2C) and always minimized the error to the impact point, regardless of whether it was real or virtual (difference from zero: p=0.903 (real), p=0.455 (virtual); One-Sample Signed Rank) with no significant difference in variance (p=0.589, Brown-Forsythe). The turns also had the same latency (p=0.297; Mann-Whitney; Fig. 2D) and kinematics (Fig. 2E: no difference in bending duration under real and VR conditions, p=0.071, Mann-Whitney). Furthermore, responses came from a large range of orientations and distances and, most importantly, were equally accurate across all distances (Fig. 2F) and turn sizes (Fig. 2G). In summary, despite the limited frame rate of the screens and despite the need to use feeders in most presentations, the turn decisions can be studied exclusively with virtual stimuli. This sets the stage for now changing the relation between initial movement and the rewarded virtual landing point.

Challenging the decision with a new rule to link input and rewarded output

The setup now allowed us to study the decision after a systematic change in the rule that connects the input and the turn the fish needs to make to reach the virtual impact points. Figure 3A introduces the model behind the new rule that we used to generate a systematic large deviation of 14.9 cm between the new VIPs and the previous ballistic VIPs. After having travelled straight throughout what was the previous decision time, the direction was deflected by 39.8° (to the left, s. Methods) and the reward was now presented at the appropriate ‘deflected’ VIP. Hence, there is still a fixed relation between the input (initial target movement on the screen) and the rewarded point on the water surface. However, the relation is now different from the one the fish had previously been using. It is important to stress, that learning to turn according to this new rule is not at all simple and would require the fish to learn non-trivial corrections (Fig. 3B): Relative to the turn they previously made for a given input constellation they would be required to add a correction that is not constant but depends sensitively on the position of the responding fish. This map of corrections is, furthermore, not the same for all input constellations but a new map as in Fig. 3B is required for each combination of input variables (as illustrated in Fig. S3). The predictions were clear for adult archerfish: should the decision use a circuit that was hardwired to ballistic movement, then the fish would never be able to turn to any other than the ballistic VIPs. The fish would still have several options to efficiently adapt to the situation, for instance by learning how to modify their subsequent approach path, by employing slower responses (not the so-called C-starts employed in the turn decision) or by waiting until the deflection had occurred to launch their turns only after that time. But they would not be able to adjust their rapid turn decisions.

The high-speed decision learns a new rule of how to connect input and output.

(A) Sketch of how we employed virtual reality to challenge the fish with a new rule of how they must turn based on initial target motion. For all objects shown, the trajectories were always straight during the time it takes the fish to make its decision but then were deflected as illustrated in the inset. Reward is then given at the corresponding new ‘deflected’ impact point but no longer at the ballistic impact point (that ballistics would predict from the initial movement). (B) Despite its apparent simplicity, learning to adjust the turns to this type of motion would require non-trivial corrections to the previous (ballistic) turning angles. These corrections depend on the fish’s initial position. Furthermore, a different such map would be needed for each set of input variables (see Fig. S3). (C, D) Median errors of the turns relative to the new and to the old ballistic (D) VIPs in successive training stages 1 to 9 (each with 60 evaluated turn decisions), with food always delivered at the new VIPs. Colored background highlights when turns were oriented towards the new (C) or to the old ballistic (D) VIPs. (E, F) Cumulative density functions (CDFs) at the stages indicated by colored circles in (C, D) to show how systematic error to the new landing point decreased (E) whereas errors to the ballistic landing points increased (F). (G) Errors made in several critical tests – from left to right: Deflection of trajectory visible and feeders present at the new VIPs (n=240), interspersed tests with only the initial straight trajectory (but not the later deflection) shown (n=68), interspersed tests (rate: 1 in 8) with also only the initial straight trajectory shown but direction of motion offset from the direction of the feeders by either 15° (n=197) or 72° (n=145) to create a clear conflict between turning to the feeder position or to the VIP derived according to the new rule. These controls show that the fish had indeed learned to base their turn decision on the initial movement. (H, I) After training the fish were able to again respond appropriately over a wide range of distances (H) and required angles of turn (I). Data are represented as median (C, D, G). ***p<0.001, **p<0.01, n.s. not significant.

We now exclusively challenged the fish with the deflected trajectories and the new rewarded VIPs over the course of several weeks. During this time, we continuously sampled occurrence and error of all turn decisions (Figs. 3C, D). Interestingly, the fish never ceased to respond with their characteristic turns (response probability remained constant at all stages, p>0.08, Pairwise Chi-square) of unchanged kinematics and never significantly increased their latency (bending and straightening phase of their C-starts both p>0.28, latency p>0.07; Kruskal-Wallis). Furthermore, their aims were systematically in error for the initial stages of training (deviation from zero error to the deflected VIPs p≤0.001, One-Sample Signed Rank) as we had predicted (Fig. 3D; deviation of aim to ballistic VIPs from zero error p>0.06; One-Sample Signed Rank; Movie S3). However, gradually the errors made to the new ‘deflected’ VIPs became smaller and, after more than 1750 presentations, the turns clearly minimized the error to the appropriate new ‘deflected’ VIPs (Figs. 3C, E, F; Movie S4) and no longer to the ballistic ones.

This surprising finding does by no means show that the decision has now been based on the new rule. For example, the fish could have learned to feed information about the deflection of the prey’s trajectory into their unfolding turns and so achieve their new final turning angles. To control for this, we interspersed tests in which the trained fish were only shown the initial linear part (without the deflection) among many presentations with the full trajectory (including the deflection). However, the errors made in the interspersed tests (n=68) were not different from those made when the full trajectories were visible (Fig. 3G). This shows that the starts were still only made using the initial motion and did not require seeing the deflection. A second explanation for the new turns would be that the fish could have learned to now take the feeder positions into account. They would then use initial movement merely as an indicator which feeder to approach. To test this, we interspersed tests in which the direction of initial target movement was always either 15° or 72° off from the feeder position and in which, additionally, exclusively the initial movement (before deflection) was shown. Moreover, in these tests feeders were offset both from the ballistic and the new VIPs. Under these conditions, the turns the fish made were still to the new deflected VIPs (differences from zero error p=0.721 at 15° (n=197) and p=0.113 at 72° (n=145); One-Sample t-tests) but not to the feeders or to the ballistic VIPs (Kruskal-Wallis: pσ0.001 at 15°; p=0.005 at 72°; Dunn’s p<0.05 for comparison of errors; (Movie S5)). Furthermore, the turns made when only the linear initial motion was visible and when the VIPs were spatially offset from the feeders did not differ in latency (p=0.686, Kruskal-Wallis) nor kinematics (bending duration: p=0.242, Kruskal-Wallis) from the turns made without offset and with the full trajectory visible (n=240).

Finally, it is important to stress that the fish had not learned to directly turn to the new VIPs just for a few restricted constellations (Fig. 3H, I). Instead, turns came from a broad range of distances and turn angles and across those had equal accuracy. So, the high-speed decisions are, surprisingly, not hardwired to a ballistic relation between the input variables and the turn the fish must make but can be reprogrammed according to a new rule

After restricted training to the new rule the decision immediately generalizes to untrained input constellations

Because all training to the new rule had occurred with targets presented at only one height level it is now possible to run generalization tests to examine the level of abstraction at which the decision was re-wired. Suppose the fish had adjusted its ballistic decision network in a way that can be formally represented as overwriting a ‘lookup table’. Such a table (or its mechanistic realization) would store the past successful turn for a given orientation and position of the fish, speed, direction, and height of moving prey. Because training was restricted and had exclusively occurred at only one height level, the previously ballistic ‘lookup table’ would have been only partially changed and predictable problems should occur in tests with targets presented at untrained heights. For these untrained heights the table would either only contain entries for ballistic motion (prediction A) or the ballistic entries might be substituted with entries that were appropriate at the training height (prediction B) during retraining. Both predictions allow us to calculate the turns the fish would have made when they first encountered targets at untrained height levels (Fig. 4A): Turns should either be to the ballistic landing point as appropriate for the novel height level (prediction A). Or, according to prediction B, the turns should be to the ‘deflected’ landing point but appropriate for the training height and not the actual height level. To test these predictions, we challenged the trained fish with movement that occurred at either higher or lower than the training height. Testing heights were chosen such that the fish would make an error of at least 6 cm relative to one of the predicted points. Figs. 4B, C show the error of the first 30 turns made when the trained fish encountered movement at the untrained height levels. The error of the turn is shown relative to the VIP that would be appropriate for the now larger (Fig. 4B) or lower (Fig. 4C) height level. This error was around zero with no tendency to change over the trials (linear regression: all p>0.1). A detailed analysis of the errors (Figs. 4D, E) showed that the turns minimized the error to the deflected landing point at the true (novel) height (difference from zero: both p>0.07; One-Sample t-test), but neither towards the ballistic landing point (prediction A; both p<0.001; One-Sample Signed Rank) nor towards the deflected landing point at the training height (prediction B; both p<0.001; One-Sample Signed Rank). Again, none of the characteristics of the turns were different in the novel situation than in the situation the fish had encountered during training (also n=30 right before these tests): latency: p=0.098, kinematics: p=0.177 (Kruskal-Wallis), variance of errors: p=0.188 (Brown-Forsythe). In other words, the new rule is represented in a way that allowed immediate generalization to the new untrained heights.

The new rule is represented in a way that allows immediate generalization.

(A) Idea for generalization tests that are possible because training to the ‘deflected’ trajectories (Fig. 3) was only for one level of target height. When faced with a larger than training height, fish would only be able to turn to the corresponding impact point P if they had acquired a more general rule of how to connect input (height, direction, speed) to the rewarding turn. Had they only substituted those input-output experiences that were no longer rewarded, then they should still use the previously rewarded input-output relations and turn to the ballistic impact point A (red, see text). Had they substituted the old input-output relation for all height levels with the trained new ones, then they should turn to point B (blue, see text) that would be appropriate for the training height. (B, C) When the fish were shown the deflected trajectories at larger (B) or lower (C) height than experienced during training, then already their first turns minimized the errors to the VIPs that are appropriate for the new rule at the untrained new height level with no indication that errors would initially be large and then decrease. (D, E) Closer analysis, using cumulative density functions (CDFs), for the errors made to the predicted points based on hypotheses introduced in (A), both for the larger (D) and for the lower (E) height (both n=30). (F) Evidence that learning the new rule had not used prior assumptions on target size. Errors of the turns made in the first 30 tests in which absolute target size was more than three-fold (13 mm) than the target size (4 mm) encountered throughout the training to the new rule. (G) CDFs of the first 30 tests with a low target speed (1.425 m/s) that the fish never encountered during training to the new rule. Again, the fish immediately chose turns to minimize the error to the predicted VIP based on the new rule but not the error to the points predicted by hypotheses A and B.

We next tested whether the fish would also generalize to new untrained target size and untrained speed levels. Fig. 4F shows that after a change in absolute target size, the trained fish still turned to the deflected landing points (differences from zero error: each p>0.4, One-Sample Signed Rank; difference in aim before and after size change p=0.764, Mann-Whitney) and the error distributions were not significantly different (difference in distributions in Fig. 4F: p=0.278, Brown-Forsythe). This shows that learning the new (deflected rule) did not involve simplifying a priori assumptions about the absolute target size. Learning also allowed immediate generalization to a lower target speed, that was not encountered during the training period. To examine this, tests with lower target speed were interspersed among presentations in which targets had the same speed as during training. Fig. 4G shows that already the first 30 turns minimized the error to the deflected VIP for the actual speed (difference from zero p=0.102; One-Sample t-test). They also did not minimize the error to the VIPs given by either ballistics (analogous to prediction A above; p<0.001; One-Sample Signed Rank) or by using predictions for the deflected VIPs that are based on the trained speed levels (analogous to B; p<0.001; One-Sample Signed Rank). Comparing latency, bending duration and the variance of errors (speed: last 30 presentations with only training speed; size: last 30 presentations with target size during training) showed that changing target size and speed also did not affect the nature of the turns. Latency and bending duration were not significantly different in the changed versus training conditions (At novel untrained speed: Latency: p=0.280, Mann-Whitney; Stage 1: p=0.076, Mann-Whitney. At novel untrained size: Latency: p=0.305, t-test; Stage 1: p=0.997, Mann-Whitney). Variance was even slightly reduced (and not increased) at the untrained target speed (p=0.021, Brown-Forsythe) and remained unchanged after the increase in target size (p=0.742, Brown-Forsythe). In summary, all tests show that the high-speed decisions had been reprogrammed in much more clever ways than predicted. Rather than independently changing the size of an unrewarding turn for each individual constellation – as would seem required because the input variables themselves are varied independently from each other – learning allowed immediate generalization and thus must have in some way captured the new rule.

The decision can conditionally operate on two different rules

We next asked whether the fish would – at any given time – only be able to use one rule for the turn they must make. To explore this, we randomly showed the trained fish two types of objects that differed in shape: Either (i) a disk that moved on the ‘deflected’ trajectory and to which they responded with turns toward the rewarded deflected VIPs, or (ii) a ‘ballistic’ object (triangle) that moved straight and for which a reward was given at the ballistic VIP (Fig. 5A). In the experiments, one of the two objects was chosen randomly, appeared in stationary position for 2 s, and then started moving. Initially, the turns made in response to the ballistic object still minimized the error to the deflected VIP (difference from zero error p=0.256; One-Sample Signed Rank; same variance as for disk p=0.343; Brown-Forsythe; first 160 responses to ballistic object) and not to the ballistic VIP point (difference to ballistic landing point: p=0.038, Mann-Whitney; Fig. 5B, C). However, after 500 additional starts in response to the ballistic objects, turns were equally in error to the deflected and the ballistic VIPs (p=0.648, Mann-Whitney). During the same period, in contrast, the turns to the moving disks remained perfectly matched to the deflected landing point (difference from zero error p=0.273; One-Sample Signed Rank; Fig. 5C). After about 300 additional exposures to the moving triangle the turns minimized the error to the ballistic impact point when the triangle was shown (error to ballistic p=0.836, error to deflected p=0.020; One-Sample Signed Rank; Fig. 5B) but continued to minimize the error to the deflected impact point whenever the disk was shown (error to ballistic p≤0.001, error to deflected p=0.084, n=187; One-Sample Signed Rank; Fig. 5C).

The high-speed decision can conditionally use two different rules of how to connect input and output.

(A) After successful training to infer the new (deflected) virtual impact points (VIPs), fish were randomly shown either the disks with the deflected trajectories (reward at deflected VIPs) or new non-disk objects that moved on straight trajectories and were rewarded at the corresponding ballistic VIPs. (B, C) Error in the turns as determined in interspersed tests that only showed the straight initial movement both for non-disk objects (triangle, B) and for the familiar disk objects (C). Objects were chosen randomly. ‘Pre’ (filled symbols) denotes baseline before fish were exposed to a total of 500 presentations of the non-disk objects with reward at their ballistic VIPs. Errors were determined in subsequent test phases. As the fish changed their turn decision to minimize the error to the non-disk VIPs (B) they continued to aim to the deflected VIPs whenever they encountered moving disks (C) showing that they had not reversed to generally using ballistics again. (D-F) Interspersed tests that only showed the same initial linear trajectory but with either a disk (n=60) or a triangle moving (n=187). Graphs on right side to illustrate choice situation as seen from above: For the same set of input variables (speed, height, direction) the fish must turn to the ballistic VIP when a triangle is moving, but to the deflected VIP when a circle is moving. Detailed analysis shows that turns were appropriate to the correct VIP with no difference in error (D), latency (E) or kinematics (bending duration) (F). (G) Individual fish were able to select the appropriate turn to the deflected VIPs when encountering moving disks and to the ballistic VIP when encountering triangles. Aims were equally appropriate and did not differ in variability. Data are represented as medians (B, C, G). n.s. not significant.

Next, we interspersed tests that only showed the initial straight paths both for the disk (n=60) and for the triangle (n=187). So, picking either the ‘ballistic’ or the ‘deflected’ rule to choose the appropriate turn for a given set of input variables could now exclusively be based on target shape. Again, the turns were toward the ballistic landing point when the triangle was shown (difference from ballistic p=0.075, from deflected p<0.001; One-Sample Signed Rank) and to the deflected landing point when the disk was shown (difference from ballistic p=0.008, from deflected p=0.412; One-Sample Signed Rank) (Fig. 5D). Also, latency (p=0.189, Mann-Whitney; Fig. 5E) and kinematics of the turns (bending duration; p=0.984; Mann-Whitney; Fig. 5F) were the same in the responses to triangles and disks. In other words, the nature of the turn decision was the same, but object shape determined which rule was used to determine the appropriate turn given the input data.

In principle these results could have arisen rather simply: some fish could have only responded to disks and others only to triangles, so that no single given individual would have to be able to handle both rules. Analyzing the contributions of the individuals showed that this was not the case, and that individual fish could respond appropriately to both conditions (Fig. 5G). Each of the three individual fish that contributed most starts had minimized the error to the appropriate VIP (difference from zero error always p>0.1; One-Sample Signed Rank) and the variability of the errors in applying the two rules did also not differ in any of the individuals (p=0.966; Brown-Forsythe). This shows clearly that individual archerfish are indeed capable of simultaneously operating on two different rules – one for each type of prey – of how to connect the input with the turn the fish must make.

A shape cue is needed to select which rule to use but is not needed prior to target movement

The previous series of experiments presented target shape as a potential prior that would inform the decision circuit as to which of the two rules will have to be used when target movement starts. To test whether it was necessary to instruct the circuit prior to the onset of motion, we interspersed tests in which the target object (disk or triangle) revealed its shape only at motion onset (Fig. 6A). In each of these interspersed tests, first a neutral shape (a cross) appeared, remained stationary for 2s, and then was exchanged randomly with either the moving triangle or the moving disk. Moreover, only the straight initial path was shown so that shape after motion onset was the only cue for the decision on whether to choose a turn based on the ‘ballistic’ or the ‘deflected’ rule (Movies S6, S7). Under these conditions, the decisions remained appropriate to the ballistic (triangle; n=109 starts) or deflected (disk; n=46 starts) VIPs (difference from zero error to appropriate VIP: p=0.451 (triangle) and p=0.671 (disk); p ≤ 0.001 for the alternatives, all One-Sample Signed Rank). Furthermore, also the distribution of errors was not statistically different from those obtained with the shape cue available 2s prior to the decision (p=0.269; Brown-Forsythe). Hence the information that is needed to select which rule is to be used can also be extracted in the very brief interval (as little as 40 ms) after motion onset, together with all other decision-relevant information but does not need to be available to instruct the decision circuit before target movement starts

The clue indicating which rule to use needs not to be known before target movement starts.

Schematic illustration of the idea to examine whether shape information was required before movement information, giving the decision circuit some time to adjust which rule was subsequently to be used. Either the randomly chosen object or a neutral symbol (cross) was shown for 2s and then the chosen object (disk or triangle) started moving. In all tests, only initial straight movement was shown. Errors were the same, regardless of whether object identity was revealed before or only during the very brief decision time (see Movies S6, S7). Data are represented as medians. n.s. not significant.

Discussion

In their turn decisions, archerfish quickly observe the initial values of motion of falling prey and then turn at the ballistic landing point. When shown purely planar initial movement they still turn at the ‘virtual’ ballistic landing point, that specific point at which ballistics predicts it to impact at the water surface. Based on this property we established a paradigm that allowed us to exchange the rules of ballistics with a new one. In this paradigm, the fish can no longer use ballistics to connect the independent variables (height, speed, and direction) with the appropriate turn towards the point of reward. However, a new rule exists that would, in principle, allow the fish to connect the rewarded turn to the combination of input variables. We show here, that, against all expectations the fish learn to extract the new rule in a way that allows them immediate generalization to variable combinations (such us untrained new height levels) they have not been trained with. The reprogramming of the decision does not change its kinematic aspects or latency, showing that the high-speed decisions were indeed reprogrammed when a new relation exists that connects prey initial movement with the point of reward.

Our findings demonstrate that reflex-like speed of a decision does not rule out the involvement of learning and of cognitive aspects that are seen in much slower decisions and in behaviors shown at more than 100 times lesser speed. They thus provide a counterexample against everyday wisdom that what occurs at reflex speed must be a simple reflex and limited to fixed or mapped inputs and responses. Although we demonstrate this in a highly specialized decision that has evolved under particularly heavy constraints, it is not unlikely that similarly flexible and adaptive decisions at reflex speed might be much more widespread. They may be common in hunting contexts and employed whenever flexibility and the efficient re-programming of rapid decisions would be a major advantage. Nontrivial elements of flexibility could also be part of rapid perceptual decision making. Given the high temporal precision obtainable in urgent perceptual decision-making (5,17) using this approach to compare cognitive aspects in the ‘classic’ slower (38,39) and in rapid (3,5) perceptual decision making could be very rewarding.

Comparing cognitive aspects of high-speed decision making with those of much slower behaviors

We would like to stress that none of the cognitive aspects we describe here is in any way simpler when compared to its slower versions. The capacity to generalize efficiently has been demonstrated in the ability of archerfish to learn of how to engage moving targets and to judge their absolute size (40,41). Also, recognizing and assigning a value to different visual shapes has been demonstrated much earlier (38,42). Such aspects are also not unique to archerfish and have been described across the animal kingdom, including animals with far smaller brains (43–45). What is surprising is, of course, clearly not to find cognitive features in archerfish, but rather to find such features of similar complexity in a decision that is made at reflex-like speed on a timescale of milliseconds rather than one of minutes. Also, the complexity of the task that the fish needs to master to conditionally adjust their high-speed decisions is certainly not easier than what otherwise is considered a demanding task in many laboratory animals (e.g. 46-48). Here a much-studied task is to have an animal under condition ‘a’ respond to a stimulus ‘α’ by turning left and right to stimulus ‘α’. Under condition ‘b’, however, the animal must do the opposite and turn right to stimulus ‘α’ and left to stimulus ‘α’. In comparison, the task the decision circuit of archerfish handles in just 40 milliseconds is not only to conditionally switch between two motor outputs but to switch between two rules of how to connect a three-dimensional set of input data to a continuum of possible motor outputs.

Highspeed decision-making and circuit size

The plausible assumption that reflex-speed implies reduced complexity and flexibility (48,49) could have a similar foundation as the assertion that smaller brains have reduced cognitive capacities. Substantial progress has been made on this latter assertion, resulting in an impressive and continuously growing list of sophisticated cognitive features that were found in animals with bird-and fish brains (50–52). Most impressively, it has been demonstrated that insect brains that operate with a millionth of the neurons of a human brain can handle a wide range of cognitive tasks (43–45,53–55) that even include social or observational learning (55–57). The reflex-speed of some decisions also will likely limit the size of the underlying neuronal circuitry, allowing only a subset of the brain to be used, a subset that could be considered a ‘functional mini-brain’. The assertion that everything that is fast as a reflex must be simple might therefore suffer the same fate as did the assertion that brains with fewer neurons must be simple. In the turn decisions of archerfish, specifically, it is known that the kinematics of their turn decisions is completely equivalent to the so-called C-starts that the fish produce to escape (37). Both the turn decisions (also called predictive C-starts) and the escape C-starts are among the fastest C-start known in fish (37) and so at least the motor circuits are restricted to the well-known powerful hindbrain circuits of a comparably small number of tractable numbers of neurons, organized around the giant Mauthner neuron (59–62). Its rapidly conducting axon would be suited to initiate the turn but it would clearly have to work together with other hindbrain neurons to account for the precision and adjusted straightening speed of the fish (37).

Ecological considerations: Why did archerfish evolve flexible highspeed decision-making?

The rapid and precise turn decisions are required because of severe competition with other surface-feeding fish that are not only more numerous – and therefore likely to be closer at the landing point of the archerfish’s prey – but that are also capable of using vision to guide pre-impact turns and therefore need to be outperformed (31). We suggest that the flexibility of the archerfish’s decision was required because of scarcity of aerial prey. If the fish must catch whatever scarce prey is available and if they are in serious competition with other fish, then they must address the following problem: While small prey will be completely wetted by the archerfish’s shot and will fall ballistically (32,35), larger prey will face considerable frictional forces and – all other conditions being equal – will therefore not fall equally far. Simply using a hardwired circuitry for ballistically falling prey also for such larger prey would produce wrong initial turns which would require subsequent corrections that cost precious time and might mean losing the advantage over the competitors. The flexibility we found and the ability to operate on two types of prey that fall according to two different rules would allow archerfish to efficiently handle two types of prey that co-occur during a specific time of the year. When one is replaced by another type of prey with yet another way in which its initial movement relates to its impact point, the decision will be able to adjust, again, to the new prey type, being still able to accurately respond to the prey that remained. So, we suggest that the ecological constraints and the severe competition that archerfish face in the wild could have favored the evolution of a decision that needed to go right to the limits of combining speed, accuracy, and flexibility.

We hope that our findings, obtained in a special case, stimulate further studies to better understand whether there is a cognitive divide between slow and fast decision making. If no such divide could be found, then this would be a strong signal to exploit rapid and flexible high-speed decisions in technological applications.

Materials and methods

Animals

Experiments were conducted on a group of six adult archerfish (Toxotes chatareus). The experimental fish were randomly chosen from a larger group of (male and female) archerfish. They had been imported from Thailand (Aquarium Glaser, Rodgau, Germany) and were kept at 26°C on a 12h:12h light/dark cycle in the large experimental tank (1m x 1m x 0.6 m, filled to a height of 34 cm) in brackish water (3.5 mS cm-1). All procedures were conducted in accordance with the German Animal Welfare Act (Tierschutzgesetz) as well as the German Regulation for the protection of animals used for experimental purposes or other scientific purposes (Tierschutz-Versuchstierordnung) and were approved by the government of Lower Franconia (Regierung Unterfranken, Würzburg, Germany).

General setup

To facilitate recording the fish from below – through the transparent bottom of the tank – together with the visual stimuli shown on the LCD monitor above, the tank was placed on steel frame stands, 110 cm above ground. A construction made from 3 cm aluminum profiles (item Industrietechnik GmbH, Germany) was used to hold the monitor on top of the tank as well as to position the feeders that were employed to reward the fish. To increase the contrast in the recordings of the fish, three conventional 500 W halogen floodlights, located around the tank, illuminated a square of white cloth (2x2 m) that was placed parallel to the water surface and 106 cm above it.

Monitors and stimuli

To present visual stimuli a 27’’ LCD monitor (Samsung SyncMaster S27A750D, 120 fps, max. intensity 300 cd/m2) was used. A glass plate in front of the screen protected it against shots of brackish water fired by the fish. All objects used in this study had a Michelsen contrast of 0.96, as determined with a luminance meter (LS-110 Konica Minolta Sensing Europe) under the conditions of the actual experiments. The surface of the LCD monitor was parallel to the water surface, h = 28.8 cm above it and its center was directly above the tank’s center. Stimuli were generated using Apple Motion 5.4.7 and stored as 120 fps FullHD ProRes movies. We presented the movies with the presentation mode of Quicktime Player 10.4 (Apple Inc.) running on a MacPro 5.1 under MacOS 10.11. An AMD 4GB Radeon RX 480 graphic card produced the monitor signal and was connected via DisplayPort with the LCD monitor. Unless stated otherwise each presentation started with an object appearing in the center of the screen. After 2s the object starts moving in one of four pre-assigned, randomly chosen directions. Objects displayed were either disks of 4 mm diameter, disks of 13 mm diameter or equilateral triangles (height 13 mm). Objects that mimicked ballistically falling motion moved at a horizontal speed vh of vh = 1.775 m s-1 and their virtual landing occurred 243 ms later, after a (virtual) path of 43 cm. On the screen, the movement of the ballistic objects was visible for 116 ms. Objects on the ‘deflected’ trajectories moved at the same horizontal speed but changed the direction of motion after being visible for 116 ms (or 20.5 cm) at a (virtual) time of 118.7 ms (i.e. after an initial straight path of dr = 21.1 cm) by an angle ɣ = 39.8° to the left (see Fig. 3A). They were then visible for a further 58 ms. The systematic change in direction causes the virtual impact point, the one inferred from initial motion, to be systematically offset by o = 2*(vh*sqrt(2h/g) - dr) * sin(ɣ/2) where h is the height at motion onset, g is the gravitational constant and dr is the distance travelled prior to the change in direction. In our experiments we selected conditions so that the offset was large enough (14.9 cm) to allow us detecting any errors in the bearing of the C-starts. In critical tests only the straight initial movement (i.e. the initial path of 20.5 cm) was shown both for ballistic and deflected trajectories.

Stimulus-coupled feeders

Unless stated otherwise 4 feeders were typically present. During training they were located at the places of possible virtual impact points. In many tests, the virtual impact points were systematically offset from the feeders by changing the direction of motion shown on the screen. This way the virtual landing points differed both from the position of the feeders as well as from the position relative to landmarks the fish might have been using during training. The feeders were custom-made (Fig S2A) and consisted of a tube (2 cm inner diameter, 44 cm long) with a sliding plate inserted 20 cm below its upper end to block the passage of a food pellet (one half of a Sera Cichlid stick). Each feeder was equipped with an electromagnet that can retract the slider to allow the passage of the pellet. All feeders were pre-loaded at random intervals before the tests, using a funnel (6 cm diameter) on the top side of each feeder (Fig S2A, B). Which of the four electromagnets was activated at which time was defined by trigger pules on the channels of the audio track of the movies shown on the LCD monitor so as to achieve an impact of the pellet at the virtual impact point and after the expected time T=sqrt(2h/g) of falling (h=initial height, g=gravitational acceleration). Four audio-channels were used to control which of the four electromagnets was to be activated. To achieve the required number of channels we connected an AV-receiver (Sony STR-DH520) to the computer via HDMI. Four output channels of the AV-receiver delivered the trigger signals to a custom-made amplifier which activated the electromagnets by means of a 200 ms pulse (12 V/300 mA). An additional audio channel triggered a pulse generator (TGP110; TTi, England) that started the high-speed video recording.

High-speed imaging

All recordings were made from below through the tanks transparent bottom (Fig S2C, D). A Fastcam MC2 color 500 (Photron, Japan) high-speed camera, operated at 500 fps (512 x 512 pixels) and equipped with a CVM0411ND f1.6/4.4-11 mm lens (Lensation, Germany) was used to monitor the complete tank. The pre-trigger option of the camera was always used and set to record 4 s before and 4 s after the trigger signal (i.e. the onset of target motion). An Apple Macbook Pro 6.2 running Microsoft Windows 7 controlled the setup via Fastcam viewer 332. Recordings were stored as Quicktime movies and contained the appearance of the object on the screen, its motion, and the impact of food delivered by one of the feeders.

Discriminating the responses of individual group members

Long-term studies of the archerfish start decisions require the competition in a group (27) and cannot be carried out in fish kept individually. It is therefore important to find ways to identify the individual group members that contributed a decision. Here, individuals were recognized based on the contour of their frontal body as seen from below. A template-contour was used for each individual fish that was drawn starting at one pectoral fin, over the snout and ending at the other pectoral fin. Respective images of the high-speed recordings were extracted using Image J (63) just when the straightening phase (so-called stage 2) of the so-called C-start (Fig. S1A) that leads to the turn was terminated. Note that the turns are made right below the water surface and thus in a fixed distance from the camera. The contours taken at the end of stage 2 were specific for each individual fish in the experimental group. Discrimination performed this way coincided with one obtained independently by additionally recording the responses from the side (Photron Fastcam MC2 color 2K, equipped with same type of lens and also operated at 500 fps, view orthogonal to the front side and 50 cm away from it) in which the individuals could be identified by means of their individual stripe-patterns.

Natural stimuli used to test the virtual reality approach

A non-transparent platform (4.5 cm in diameter) was mounted above the center of the tank. This allowed to blow food pellets off the platform from a similar initial height as the virtual objects shown on the screen. Movement was started by an airflow (as in 27,35) and pellets fell ballistically toward the water surface.

Quantitative analysis of the turn responses

Image J 1.50 (63) was used to analyze the recorded videos. To ensure that the decisions could not be influenced by those of other fish, we exclusively analyzed the turn of the fish that responded first in a given test. Latency was the time between motion onset and onset of the C-start. Error in aim relative to a point of interest (e.g., the virtual impact point) was determined from the last frame at the end of the straightening phase (stage 2) of the C-start, i.e., before the fish actually starts its approach trajectory (Fig. S1A). Orientation of the fish’s rigid front end was determined from the tip of the snout and the point midway between the two pectoral fins. If a line drawn along this orientation did cut a second line drawn between the start of motion and the point of interest (e.g., the virtual impact point), both projected to the water surface4, then the error was defined negative, otherwise it was defined positive (Fig. S1B). Additionally, we determined kinematic aspects of the C-starts, such as the duration of its first bending stage (stage 1 of the C-start), the mean angular speed (i.e., turn angle / turn duration), and the initial orientation and distance the fish had before the onset of their C-start. Only C-starts that were completed before prey impact were included in our analysis, to ensure that the C-start could not have used information from the impact of the food pellets. Furthermore, only C-starts were included in which the subsequent approach path of the responding fish was not blocked by another fish. When aims had to be evaluated relative to two points of interest (e.g., virtual landing point and feeder position) then (i) both possible approach paths had to be free and not blocked by any group member and (ii) the responding fish had not to be oriented in a line with both points. Analysis of the coordinates and frame numbers determined with ImageJ was carried out using Excel (14.0 & 15.11).

Statistics

Statistics were run on GraphPadPrism 9.5. (GraphPad Software, USA). Tests referred to in the main text were as follows: Normality of data was assayed using Shapiro–Wilk tests. For single data sets (error of aim), difference from zero was checked using One-Sample t-tests (parametric) or One-Sample Signed Rank tests (non-parametric). Comparisons of two data sets (latency, duration of stage 1, error of aim) were performed with unpaired t-tests or Mann-Whitney tests. Comparisons of three or more data sets (latency, duration of stage 1, error of aim) were performed with Kruskal– Wallis One-Way ANOVA on ranks, post hoc tested with Dunn’s method (non-parametric). Equal variance between data sets (error of aim) was checked using Brown-Forsythe tests (non-parametric). Stability of aim was tested using linear regression. Tests for changes in C-start probability employed pairwise comparisons with χ2 tests. All respective tests were performed two-sided. ***p<0.001, **p<0.01, n.s. not significant. All data are available in the main text or the supplemental information.

eLife Assessment

This valuable study investigates prey capture by archer fish, showing that even though the visuomotor behavior unfolds very rapidly (within 40-70 ms), it is not hardwired; it can adapt to different simulated physics and different prey shapes. Although there was agreement that the model system, experimental design, and main hypothesis are certainly interesting, opinions were divided on whether the evidence supporting the central claims is incomplete. A more rigorous definition and assessment of "reflex speed", more detailed evidence of stimulus control, and a more detailed analysis of individual subjects could potentially increase confidence in the main conclusions.

Reviewer #1 (Public review):

Summary:

The authors test whether the archerfish can modulate the fast response to a falling target. By manipulating the trajectory of the target, they claim that the fish can modulate the fast response. While it is clear from the result that the fish can modulate the fast response, the experimental support for the argument that the fish can do it for a reflex-like behavior is inadequate.

Strengths:

Overall, the question that the authors raised in the manuscript is interesting.

Weaknesses:

(1) The argument that the fish can modulate reflex-like behavior relies on the claim that the archerfish makes the decision in 40 ms. There is little support for the 40 ms reaction time. The reaction time for the same behavior in Schlegel 2008, is 60-70 ms, and in Tsvilling 2012 about 75 ms, if we take the half height of the maximum as the estimated reaction time in both cases. If we take the peak (or average) of the distribution as an estimation of reaction time, the reaction time is even longer. This number is critical for the analysis the authors perform since if the reaction time is longer, maybe this is not a reflex as claimed. In addition, mentioning the 40 ms in the abstract is overselling the result. The title is also not supported by the results.

(2) A critical technical issue of the stimulus delivery is not clear. The frame rate is 120 FPS and the target horizontal speed can be up to 1.775 m/s. This produces a target jumping on the screen 15 mm in each frame. This is not a continuous motion. Thus, the similarity between the natural system where the target experiences ballistic trajectory and the experiment here is not clear. Ideally, another type of stimulus delivery system is needed for a project of this kind that requires fast-moving targets (e.g. Reiser, J. Neurosci.Meth. 2008). In addition, the screen is rectangular and not circular, so in some directions, the target vanishes earlier than others. It must produce a bias in the fish response but there is no analysis of this type.

(3) The results here rely on the ability to measure the error of response in the case of a virtual experiment. It is not clear how this is done since the virtual target does not fall. How do the authors validate that the fish indeed perceives the virtual target as the falling target? Since the deflection is at a later stage of the virtual trajectory, it is not clear what is the actual physics that governs the world of the experiment. Overall, the experimental setup is not well designed.

Reviewer #2 (Public review):

Summary:

This manuscript studies prey capture by archer fish, which observe the initial values of motion of aerial prey they made fall by spitting on them, and then rapidly turn to reach the ballistic landing point on the water surface. The question raised by the article is whether this incredibly fast decision-making process is hardwired and thus unmodifiable or can be adjusted by experience to follow a new rule, namely that the landing point is deflected from a certain amount of the expected ballistic landing point. The results show that the fish learn the new rule and use it afterward in a variety of novel situations that include height, side, and speed of the prey, and which preserve the speed of the fish's decision. Moreover, a remarkable finding presented in this work is the fact that fish that have learned to use the new rule can relearn to use the ballistic landing point for an object based on its shape (a triangle) while keeping simultaneously the 'deflected rule' for an object differing in shape (a disc); in other words, fish can master simultaneously two decision-making rules based on the different shape of objects.

Strengths:

The manuscript relies on a sophisticated and clever experimental design that allows changing the apparent landing point of a virtual prey using a virtual reality system. Several robust controls are provided to demonstrate the reliability and usefulness of the experimental setup.

Overall, I very much like the idea conveyed by the authors that even stimuli triggering apparently hardwired responses can be relearned in order to be associated with a different response, thus showing the impressive flexibility of circuits that are sometimes considered mediating pure reflexive responses. This is the case - as an additional example - of the main component of the Nasanov pheromone of bees (geraniol), which triggers immediate reflexive attraction and appetitive responses, and which can, nevertheless, be learned by bees in association with an electric shock so that bees end up exhibiting avoidance and the aversive response of sting extension to this odorant (1), which is a fully unnatural situation, and which shows that associative aversive learning is strong enough to override preprogrammed responding, thus reflecting an impressive behavioral flexibility.

Weaknesses:

As a general remark, there is some information that I missed and that is mandatory in the analysis of behavioral changes.

Firstly, the variability in the performances displayed. The authors mentioned that the results reported come from 6 fish (which is a low sample size). How were the individual performances in terms of consistency? Were all fish equally good in adjusting/learning the new rule? How did errors vary according to individual identity? It seems to me that this kind of information should be available as the authors reported that individual fish could be recognized and tracked (see lines 620-635) and is essential for appreciating the flexibility of the system under study.

Secondly, the speed of the learning process is not properly explained. Admittedly, fish learn in an impressive way the new rule and even two rules simultaneously; yet, how long did they need to achieve this? In the article, Figure 2 mentions that at least 6 training stages (each defined as a block of 60 evaluated turn decisions, which actually shows that the standard term 'Training Block' would be more appropriate) were required for the fish to learn the 'deflected rule'. While this means 360 trials (turning starts), I was left with the question of how long this process lasted. How many hours, days, and weeks were needed for the fish to learn? And as mentioned above, were all fish equally fast in learning? I would appreciate explaining this very important point because learning dynamics is relevant to understanding the flexibility of the system.

Reference:

(1) Roussel, E., Padie, S. & Giurfa, M. Aversive learning overcomes appetitive innate responding in honeybees. Anim Cogn 15, 135-141, doi:10.1007/s10071-011-0426-1 (2012).

Author response:

The following is the authors’ response to the original reviews

eLife Assessment

This valuable study investigates prey capture by archer fish, showing that even though the visuomotor behavior unfolds very rapidly (within 40-70 ms), it is not hardwired; it can adapt to different simulated physics and different prey shapes. Although there was agreement that the model system, experimental design, and main hypothesis are certainly interesting, opinions were divided on whether the evidence supporting the central claims is incomplete. A more rigorous definition and assessment of "reflex speed", more detailed evidence of stimulus control, and a more detailed analysis of individual subjects could potentially increase confidence in the main conclusions.

Thank you very much. There are several points that we had to absolutely make sure that they are very well understood. (1) Explaining in the best possible way the experiment with a fly sliding on top of a glass plate. Here, the virtual ballistic landing point can be calculated using simple high school physics. It turns out that this is where the fish turn to – even though the fly is not falling at all. Once this is understood it becomes clear that we can precisely measure latency and accuracy of the C-start turns. In the new version we explain this essential aspect in more detail and add an extra Figure (new Figure 2). This may, perhaps, help readers to notice this important background (previously covered in Fig. 1C). (2) The full experimental evidence that the VR method works is presented in more detail and all measurements necessary will be clear after the new Figure 2. They will however not be clear if this Figure is ignored. (3) We have rewritten the manuscript to make it easier to understand what we wanted to show, why we needed VR to proceed and why the archerfish highspeed decision lent itself so readily to tackle the problem. (4) We emphasize the importance of speed-accuracy tradeoffs in standard decision-making and also include data on the absence of such a relation in the archerfish highspeed decisions.

So, in summary, we have emphasized what we wanted to show and what we did not want to show, we have rewritten the text to make it easier for future readers and we have tried to add more guidance to the figures. We do hope very much that the beauty of the quite unexpected findings is more easily visible to those who take the trouble of actually reading the paper.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The authors test whether the archerfish can modulate the fast response to a falling target. By manipulating the trajectory of the target, they claim that the fish can modulate the fast response. While it is clear from the result that the fish can modulate the fast response, the experimental support for the argument that the fish can do it for a reflex-like behavior is inadequate.

Please note that we have not simply tested whether archerfish can 'modulate the fast response'. We quantitatively test specific hypotheses on the rules used by the fish. For this the accuracy of the decisions is analyzed with respect to specific points that can be calculated precisely in each of the experiments. These points are shown on the figures and in the movies that were meant to illustrate this important aspect. We had to make sure that the way we calculate the predicted point(s) is made as clear as possible in the text. We added more text and separated the fundamentally important aspects in a separate Figure 2 to make it more difficult to overlook the fundamental aspects that lay the foundation for everything that follows.

Strengths:

Overall, the question that the authors raised in the manuscript is interesting.

Thank you and we do hope very much that, with our revision, you will see the beauty of the findings.

Weaknesses:

(1) The argument that the fish can modulate reflex-like behavior relies on the claim that the archerfish makes the decision in 40 ms. There is little support for the 40 ms reaction time. The reaction time for the same behavior in Schlegel 2008, is 6070 ms, and in Tsvilling 2012 about 75 ms, if we take the half height of the maximum as the estimated reaction time in both cases. If we take the peak (or average) of the distribution as an estimation of reaction time, the reaction time is even longer. This number is critical for the analysis the authors perform since if the reaction time is longer, maybe this is not a reflex as claimed. In addition, mentioning the 40 ms in the abstract is overselling the result. The title is also not supported by the results.

Although the minimum latency is indeed 40 ms (it can be slightly less: e.g., see the evidence in the paper, for instance the plots in the new Fig. 4) the paper's statements are not dependent on a specific number. Even if minimum latency was 100 ms (which it is not) the speed of the response and the absence of a speedaccuracy relation (now shown directly in Fig. 4) is what is of importance. To show this we have completely rewritten large parts of the manuscript.

(2) A critical technical issue of the stimulus delivery is not clear. The frame rate is 120 FPS and the target horizontal speed can be up to 1.775 m/s. This produces a target jumping on the screen 15 mm in each frame. This is not a continuous motion. Thus, the similarity between the natural system where the target experiences ballistic trajectory and the experiment here is not clear. Ideally, another type of stimulus delivery system is needed for a project of this kind that requires fast-moving targets (e.g. Reiser, J. Neurosci.Meth. 2008). In addition, the screen is rectangular and not circular, so in some directions, the target vanishes earlier than others. It must produce a bias in the fish response but there is no analysis of this type.

Please note that the new Fig. 3 (former Fig. 2) reports all the evidence that is needed to just show this and in a way that could in no way have been better. We have rewritten the text to explain what needs to be shown experimentally in order to be able to proceed, what critical tests were done and what results were obtained. We also add a short comment on another unsuccessful attempt that we have tried before.

(3) The results here rely on the ability to measure the error of response in the case of a virtual experiment. It is not clear how this is done since the virtual target does not fall. How do the authors validate that the fish indeed perceives the virtual target as the falling target? Since the deflection is at a later stage of the virtual trajectory, it is not clear what is the actual physics that governs the world of the experiment. Overall, the experimental setup is not well designed.

Understanding this aspect is essential. If the glass plate experiment is not thoroughly understood (new Fig. 2 with new text to emphasize that this is absolutely essential) nothing that follows makes any sense, including what is meant by the statement that the decision could be hardwired to ballistic motion.

Reviewer #2 (Public review):

Summary:

This manuscript studies prey capture by archer fish, which observe the initial values of motion of aerial prey they made fall by spitting on them, and then rapidly turn to reach the ballistic landing point on the water surface. The question raised by the article is whether this incredibly fast decision-making process is hardwired and thus unmodifiable or can be adjusted by experience to follow a new rule, namely that the landing point is deflected from a certain amount of the expected ballistic landing point. The results show that the fish learn the new rule and use it afterward in a variety of novel situations that include height, side, and speed of the prey, and which preserve the speed of the fish's decision. Moreover, a remarkable finding presented in this work is the fact that fish that have learned to use the new rule can relearn to use the ballistic landing point for an object based on its shape (a triangle) while keeping simultaneously the 'deflected rule' for an object differing in shape (a disc); in other words, fish can master simultaneously two decisionmaking rules based on the different shape of objects.

Strengths:

The manuscript relies on a sophisticated and clever experimental design that allows changing the apparent landing point of a virtual prey using a virtual reality system. Several robust controls are provided to demonstrate the reliability and usefulness of the experimental setup.

Overall, I very much like the idea conveyed by the authors that even stimuli triggering apparently hardwired responses can be relearned in order to be associated with a different response, thus showing the impressive flexibility of circuits that are sometimes considered mediating pure reflexive responses.

Thank you so much for this precise assessment of what we have shown!

This is the case - as an additional example - of the main component of the Nasanov pheromone of bees (geraniol), which triggers immediate reflexive attraction and appetitive responses, and which can, nevertheless, be learned by bees in association with an electric shock so that bees end up exhibiting avoidance and the aversive response of sting extension to this odorant (1), which is a fully unnatural situation, and which shows that associative aversive learning is strong enough to override preprogrammed responding, thus reflecting an impressive behavioral flexibility.

That's very interesting, thanks and we are very happy to mention this important study in the revised version.

Weaknesses:

As a general remark, there is some information that I missed and that is mandatory in the analysis of behavioral changes.

Firstly, the variability in the performances displayed. The authors mentioned that the results reported come from 6 fish (which is a low sample size). How were the individual performances in terms of consistency? Were all fish equally good in adjusting/learning the new rule? How did errors vary according to individual identity? It seems to me that this kind of information should be available as the authors reported that individual fish could be recognized and tracked (see lines 620-635) and is essential for appreciating the flexibility of the system under study.

Secondly, the speed of the learning process is not properly explained. Admittedly, fish learn in an impressive way the new rule and even two rules simultaneously; yet, how long did they need to achieve this? In the article, Figure 2 mentions that at least 6 training stages (each defined as a block of 60 evaluated turn decisions, which actually shows that the standard term 'Training Block' would be more appropriate) were required for the fish to learn the 'deflected rule'. While this means 360 trials (turning starts), I was left with the question of how long this process lasted. How many hours, days, and weeks were needed for the fish to learn? And as mentioned above, were all fish equally fast in learning? I would appreciate explaining this very important point because learning dynamics is relevant to understanding the flexibility of the system.

First, it is very important to keep the question in mind that we wanted to clarify: Does the system have the potential to re-tune the decisions to other non-ballistic relations between the input variables and the output? This would have been established if one fish was found capable of doing that. We have rewritten the introduction and discussion to specifically say what our aim was. We feel that the paper is already extremely long and difficult to understand (even after we tried very hard in this revision to explain everything in detail and as good as we could), requires the establishment of a method whose success was really unexpected and finding a degree of plasticity that we did not expect at all. We also have added a section in the discussion stating what we can, and we cannot say given the number of fish examined. For instance, we do not know if there are differences in the speed at which the different individuals mastered the new rules and if social learning could play a role to speed up the acquisition. That is a brilliant idea and we are very interested in checking this - but we wanted to stick with the (quite ambitious) goal of the present study.

Reference:

(1) Roussel, E., Padie, S. & Giurfa, M. Aversive learning overcomes appetitive innate responding in honeybees. Anim Cogn 15, 135-141, doi:10.1007/s10071011-0426-1 (2012).

Thanks for this reference!

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Minor comments:

(1) What is the difference between Reinel, J. Exp. Bio. 2016 and the current study?

Clearly in that study all objects were strictly falling ballistically, and latency and accuracy of the turn decisions were determined when the initial motion was not only horizontal but had an additional vertical component of speed. The question of that study was if the need to account to an additional variable (vertical speed) in the decision would affect its latency or accuracy. The study showed that also then archerfish rapidly turn to the later impact point. It also showed that accuracy and latency were not changed by the added degree of freedom.

(2) How do Figures 2 F and G demonstrate that an accurate start is possible?

See above.

(3) Figure 4 is hard to follow, it is not clear what is presented and how it supports the claim that the new rule is represented in a way that allows immediate generalization.

Yes, this is not at all an easy experiment. Briefly, fish were re-trained at only one height level and then are tested at other levels. The strategy is as in the experiments Schuster et al. 2004 Current Biology, Vol. 14, 1565–1568, Figure 5. We have changed text and Figure (new Figure 5) to show how the predictions were reached.

Reviewer #2 (Recommendations for the authors):

Minor remarks

Lines 88-90: I was surprised to see that in this section, the authors did not mention the speed-accuracy trade-off off which has inspired numerous experiments in animal behavior (1). This could be used to back their point, namely, that speed comes with an apparent cost of a loss in accuracy.

Yes, that is a crucial aspect that was completely missing even though it demonstrates a key aspect of 'standard' versus some 'highspeed' decisions! We definitely had to include it and also to show, directly under the conditions of our present experiments (in the new Fig. 4) the absence of a significant speedaccuracy relation for the archerfish highspeed decisions! Thank you very much for emphasizing this crucial aspect!

Lines 182-184: Specify that this situation corresponds to the hatched bar in Figure (this can be specified in the caption of the figure, where the bar is not mentioned).

Thanks!

Lines 187-188: here and elsewhere (e.g. lines 224-225, etc), the error made by the fish is presented in cm (see Figure 2 where the inset shows how the error was computed). I wonder if it would not be more appropriate to present it in terms of the angular difference between the trajectory made by the fish and the food delivery location.

Angles could also be used, but because of the large variation in initial distances (that we wanted to make sure that the fish had to capture a rule, allowing them to respond from various distances) another measure was used that we found somehow more natural: it is simply how close a fish would get to the landing point if it continued in the direction assumed after the turn. Although we describe how we defined accuracy we did not discuss why this measure was used in this (and many previous studies). We are very happy to add this. Please also note that running all tests based on angular errors (which we also have done throughout to ensure that the conclusions are independent on an arbitrary measure of the error) leads to no different conclusion. We have added a brief explanation in the text and in the new Fig. 2.

Lines 299-323: Is it my impression or did fish have more trouble in generalizing their learned rule to the condition untrained larger height (see for instance red curves in Figures 4 D, E, G)? Could the authors elaborate on this point?

We changed the code to make this more clear. The red curves (before marked A to highlight impact point option A) correspond to the errors to the ballistic impact point without deflection, so what would have to be compared are the black curves (marked P to highlight the virtual impact point that should be chosen had the fish immediately generated to the untrained conditions). We have rewritten the text and the labels in the Figure (now Figure 5) to illustrate the predictions and to name them in more helpful ways and so that they can't be confused with panel labels. At any rate, what needs to be compared, to check the idea, are the black curves, and these are not statistically different between both heights (p=0.525, Mann-Whitney). Interestingly, none of the black curves from all panels (D-G) differ (p>0.3).

Line 559: if we are speaking here about luminance contrast, it should read 'Michelson Contrast' rather than 'Michelsen Contrast'.

Absolutely, thanks!

References

(1) Chittka, L., Skorupski, P. & Raine, N. E. Speed-accuracy tradeoffs in animal decision making. Trends Ecol Evol 24, 400-407, doi:10.1016/j.tree.2009.02.010 (2009).

An excellent paper that helps to stress our main question

Version 2

Abstract

Abstract

It is widely accepted that more time and information yield better decisions. However, some decisions manage to combine speed and accuracy in an unusual way. Potentially their trick could be to use simplifying heuristics that works well for the most common condition but would lack flexibility otherwise. Here we describe an unexpected level of flexibility in a complex highspeed decision that is made faster than an Olympic sprinter can respond to the start gun. In this decision, archerfish observe the initial speed, direction, and height of falling prey and then use these initial values to turn right towards were ballistically falling prey would later land. To analyze the limits in flexibility of this highspeed decision we first developed and critically tested a system that allowed us to replace the usual ballistic relation between initial prey motion and the expected landing point with another deterministic rule. We discovered that, surprisingly, adult fish could reprogram their highspeed decision to the new rule. Moreover, after reprogramming their decision fish were immediately able to generalize their decision to novel untrained settings, showing a remarkable degree of abstraction in how the decision circuit represented the novel rule. The decision circuit is even capable of simultaneously using two distinct sets of rules, one for each of two visually distinct objects. The flexibility and level of cognition are unexpected for a decision that lacks a speed-accuracy tradeoff and is made in less than 100 ms. Our findings demonstrate the enormous potential highspeed decision making can have and strongly suggest that we presently underappreciate this form of decision making.

Introduction

It is common wisdom that the quality of our decisions depends on how much time and effort we devote to them. The expected tradeoff between the speed of a decision and the accuracy that can be attained is very well documented not only in humans but for many decisions across many species from slime mold to primates (1). A class of very rapid decisions appear, however, to deviate from this rule (e.g. 2-8). In these decisions, accuracy does not increase when more time is available. For example, rats can reliably categorize odors based on just one sniff (about 300 ms), and more sampling time leads to no further improvement. This is not due to the simplicity of the task and holds equally for both simple and complex odor-discrimination tasks (4,9,10). Archerfish, the subject, of the present study, monitor the speed, direction and height of falling prey and rapidly turn towards the corresponding ballistic impact point (11,12). In this decision, no relation was found between the average response latency and the accuracy of the turns (11). Humans can reliably classify natural visual images after surprisingly brief glimpses (of only 20 to 40 ms, see 2,13-19) and more recent approaches demonstrated that the perceptual parts of human decision making can often be much faster than previously thought (6,20). Such rapid decisions are certainly common in sports. For examples, hitters in baseball have only 400 ms to move their bat to the optimal strike point, a movement that can only be based on about 70 ms of visual processing (21–23). Finally, even slow, and probably complex human decisions, such as which product to buy in a store, can sometimes deviate from the expected speed-accuracy tradeoff and be superior when made with lesser time and lesser information (7, 8, 24–26).

Despite these deviations, there is of course no doubt that most decisions do show speed-accuracy tradeoffs. But why is this type of decision much more common and why are not simply all decisions both fast and accurate? One reason could be that the decisions that lack a speed-accuracy tradeoff manage to be so fast because they profit from using simplifying heuristics (7,8). As we illustrate with a specific example below this allows being fast and accurate for specific conditions that enormously reduce the complexity of the problem. They would, however, then not be useful under more variable conditions und could lack the flexibility to adjust to them. In other words, the flexibility to cope with changing conditions could be what distinguishes the more common slower decisions (with speed-accuracy tradeoffs) and the more unusual highspeed decisions. The evidence that is presently available – from the movement of baseball outfielders and frisbee-catching dogs (27,28) to the chase of hoverflies (29) – seems to support this view (also see 7,8). The decision made by a hoverfly male that starts to intercept a passing female nicely illustrates this (Fig. 1A). As soon as a female (traveling at speed v) comes into sighting distance (D), the male rapidly decides on a turn angle that, if it constantly accelerates at its maximum rate (a), leads it right to the later point of interception. An elegant series of experiments showed that this very rapid and accurate decision is based on hardwired assumptions about the speed v and size (and hence detection distance D) of a typical female hoverfly and about the male’s maximum rate a of acceleration. These variables appear all to be genetically imprinted in the neural circuits that govern the male turn decision. Expectedly, challenging the males with objects that deviated from these assumptions, caused them to make robust errors that could be predicted based on a simple heuristic model that the fly appears to use (29). The turn decisions of male hoverflies are thus an excellent example of a very fast and accurate decision that lacks flexibility but has evolved to work specifically for a restricted set of conditions, which is perfect for this specific behavioral context (29).

Two highspeed decisions used by animals to catch objects.

(A) Male hoverflies use a rapid open-loop strategy to catch passing females: When sighting a female that moves at speed ‘v’, at sighting distance ‘D’ from the male, then the male rapidly turns so that a constant acceleration ‘a’ will automatically make him arrive at the point of interception at the appropriate time. In this decision, the male takes information only once and uses hardwired circuitry that assumes fixed values for D, v, and a. While this simplifies the complex interception considerably it would make it impossible to intercept targets with other values of D, v, or a (29). (B) Archerfish down aerial prey with shots of water and also use an open-loop strategy to arrive simultaneously with their prey as soon as it hits the water surface. The responding fish, shooters, and bystanders alike, sample visual information on the speed, height and direction of the initial falling motion and turn right towards where ballistically falling prey would later land. These turns are part of a maneuver, called C-start (34,40), that also lends the fish the level of speed that, when kept, make it arrive just in time. Depending on visual contrast, water temperature, height of prey and the distance from the future landing point, fully accurate turns can be initiated in as little as 40 ms (11). In contrast to the conditions in the male hoverfly, the ecological constraints in archerfish do not allow similar simplifying assumptions and archerfish can respond appropriately from a large range of distances and initial orientation and when encountering targets of unusual size or speed (as will be shown below).

Here we examine the degree of flexibility in another highspeed decision that lacks a speed-accuracy tradeoff and in which the ecological context could potentially make more flexibility advantageous. This decision is an essential part of the hunting behavior of archerfish, in which the fish shoot down aerial prey with jets of water (30-32; Fig. 1B). Once a prey insect is successfully dislodged, shooters and bystander archerfish have only the little time that remains till prey hits the water surface to turn and dash off to be at the impact point of prey just when it arrives there. In the wild, missing this point or not being there in time, would typically mean losing prey to the more numerous competing other surface-feeding fish (33). The solution archerfish have evolved to secure their prey and to outperform their competitors is to watch the initial movement of dislodged prey and then quickly and accurately select the appropriate turn and take-off to be at the later landing point at the right time (e.g., 11,12,34; Fig. 1B). The so-called predictive C-start decisions of archerfish are among the fastest C-starts known in fish and are initiated much faster than an Olympic sprinter can respond to the start gun (35; see Movie S1 for an appreciation of the decision’s speed).

An essential feature of the decisions – and the basis for the present study – is introduced in Fig. 2. It was discovered in experiment in which archerfish typically faced real ballistically falling prey but were occasionally challenged with prey that was prevented from falling (and only slid on top of a glass plate; Fig. 2A; 12). When many such interspersed rare occasions were later analyzed it turned out that fish still responded, although nothing fell. Moreover, their response latency – measured between the onset of prey motion and the onset of the turns – was the same as in most tests in which real prey was falling (12). Given that there never was an actual later impact point, where did the fish turn in the experiments with prevented falling? This became clear when the virtual ballistic impact points were calculated for each individual test made with sliding flies. Using simple high school physics (and confirming that the prey fell accordingly, 12) allowed to calculate the virtual landing point for each combination of initial height, speed, and direction that the sliding fly had in the interspersed tests. This combination differed from one test to the next. The accuracy of each turn decision with respect to each calculated virtual landing point could then be measured in the same way as with real falling objects (Fig. 2B). This analysis revealed that for each set of conditions the fish had turned to the appropriate virtual impact point, calculated based on ballistics, just as they would toward a real later impact point of real falling prey (12). Their turns toward virtual ballistic landing points inferred from initial prey movement imply that the fish were able to map the initial values of motion to the turn angle the fish must produce and that this mapping is appropriate for ballistically falling objects (Fig. 2B).

When prey is prevented from falling, archerfish turn to the virtual ballistic impact point.

(A) Sketch of an experiment in which archerfish are occasionally challenged with prey that is prevented from falling but that only slides horizontally on a glass plate. As long as such tests are rare, archerfish still elicit turn decisions even though the fly does not fall and will never ever come down. Remarkably the aim of these starts is towards the appropriate ‘virtual’ impact point (VIP). To show this, one uses ballistics, to calculate for each test with a sliding fly where the fly would have landed, given its initial values of motion (initial horizontal speed, direction and height). Using the results accumulated from many such tests shows that, for any combination of the initial motion values, the fish turn toward the expected ballistic landing point. The decision circuit thus maps the combination of initial motion values to a turn towards the expected ballistic landing point. (B) Sketch to show how the accuracy of turns toward the appropriate VIPs can be quantitatively assayed for each of the various combinations of sliding speed, direction, and height. An ecologically relevant measure of the turn’s accuracy is how close the chosen orientation would bring the fish to the VIP, i.e. how close following the chosen initial direction would bring the fish to the point of reward. The sketch also illustrates that this measure allows pooling responses with widely different initial conditions, i.e., wide differences in the initial distance and required turn angle for each VIP. (A, B) This experiment provides the basis of all experiments in this paper. It shows that turns can (i) be elicited by purely horizontal motion, and that (ii) these turns are towards the virtual ballistic impact points, i.e. the points predicted from initial motion, using ballistics.

In the glass plate experiments of Fig. 2 prey movement was not controlled by the fish but flies were made to slide by blowing an airstream at them. This introduces another convenient aspect of the decision that we will exploit throughout the study. The archerfish’s turn decisions exclusively use information that is sampled during a brief interval after onset of prey motion (11, 36–38). This information even includes the initial height from which prey is falling (37). In other words, all experiments available have shown that the start decision does not use or has no access to prior information. Instead, it relies exclusively on information sampled once prey starts moving.

With this background it may now become clear what makes the turn decisions of archerfish so interesting for studying the limits in flexibility in highspeed decision making. Because the turn decisions are appropriately made to the virtual ballistic landing point of prey that is not even falling, it is clear that the fish’s turn is somehow tuned to ballistically falling prey. This allows the fish to turn to the appropriate virtual landing point for each given combination of initial values of prey motion (Fig. 2). Because all archerfish studied so far in the lab and in the wild consistently showed robust turn decisions towards the inferred ballistic landing points (11, 12, 33, 36–40) we expected the decision to be hardwired to ballistics. Hardwiring the decision circuit to ballistic falling patterns could be fixed genetically and/or during the early ontogeny of the fish. A striking advantage of more flexibility is also not immediately obvious: While an initial aim to the ballistic landing point would be in error for non-ballistically falling prey, this initial error of the fish’s turn could always be corrected during the actual approach trajectory. Nevertheless, we demonstrate here that the decision is not hardwired to ballistic falling. We show that adult archerfish are capable of learning to extract and to use a novel non-ballistic rule of how to connect the three independent input variables of prey motion to the point to which they must turn to receive a reward.

Results

Critically testing a virtual reality approach for studying the flexibility of the decision

Experiments with purely virtual impact points and their precise quantitative analysis (as introduced in Fig. 2A,B) would, in principle, offer a way of testing whether the archerfish’s highspeed decisions are limited to ballistically falling prey. For any given combination of initial speed, initial height, and direction of prey movement and for any given orientation and position of the responding fish, ballistics requires the fish to turn by a specific angle. Now suppose the relation between initial prey movement and the point to which the fish must turn would be changed to no longer be defined by ballistics, but by some other rule. Then all turn decisions after the switch would be systematically in error and the key question is whether at least some fish would eventually be able to adjust their start decisions accordingly. Exploring this requires a virtual reality (VR) system. Because the fish need to be rewarded at the appropriate place and time to stay motivated over extended series of experiments (that last many months), it is not at all clear, if such a system could work without introducing additional cues. To be useful for the question at hand, the system needs to fulfill four conditions: (i) First, it must present the information with sufficient spatial and temporal resolution as needed by the decision circuit. (ii) Second, it needs to present the reward at the positions and times as required by ballistic or arbitrary other rules but without adding further information that could interfere with the decision. (iii) Third, the VR system would have to work so well that there is no statistically detectable difference between virtual and real responses. (iv) Finally, the system must keep the fish motivated for the many months of experimentation during which the fish will exclusively be facing virtual objects. Clearly, if only one of these conditions is not met, then the chosen VR system would not be useful to explore our fundamental question whether the decisions are or are not flexibly adjustable to non-ballistic relations.

The first step in our endeavor therefore was to explore and quantitatively test the suitability of potential setups. After several unsuccessful attempts – including LED-based arrangements used in earlier work (41–43) – we found a system that completely fulfilled all four crucial conditions. In the following we present evidence for that successful system (Fig. 3). The setup worked, surprisingly, with an LCD screen at a frame rate of 120 frames per second and employed custom-made automatic feeders that were non-transparent and provided food at the right time and place, coupled to the motion shown on the screen. To critically test the approach, the automatized feeders presented food at the ballistic virtual impact points and at the appropriate time. First, the setup readily and continuously elicited turns toward the calculated ballistic impact point (Fig. 3B-G). Next, a series of experiments showed, that the highly conspicuous feeders did not influence the decision at all but that the decision was still exclusively based on the initial target motion shown on the screen (Fig. 3B): Operating the feeders without showing motion on the screen failed to elicit turn decisions. Most importantly, when feeders were placed not at the ballistic virtual impact points (VIPs) but systematically offset from them, then turns were aimed at the VIPs and not at the feeders – even though this was where the fish got their reward (Fig. 3B; 15° offset, n=353; 30° offset, n=198). These tests were conducted so that the VIPs also differed relative to other potential landmarks in the tank and so they show that the fish were also not starting to positions marked by landmarks other than the feeders. In other words, even in the presence of the feeders were the decisions only based on the movement information that was presented on the screen and were not influenced by cues from the feeders or from any other landmarks in the tank.

Critical tests that directly show the suitability of the virtual reality setup that is used in the present paper.

To exclusively work with virtual stimuli, as required in the present study, a system was needed in which the fish would respond to virtual stimuli as in the natural situation. (A) Sketch of the successful setup with motion presented on an LCD screen and feeders operated to deliver food in time at the virtual impact points (VIPs), the points at which ballistically falling prey would impact, given its initial motion (see Fig. 2). (B) Evidence that the decisions are not influenced by the presence and operation of the conspicuous feeders: Same errors to VIPs with feeders present or removed. No rapid turns are elicited when feeders are operated without movement shown on the screen. In tests in which the feeders are not at the position of the VIPs but offset from them (by either 11.2 or 22.3 cm), turns minimized the error to the VIPs but not to the feeder positions. Data are represented as medians, inset recalls how errors were defined (see Figs. 2 and S1). (C-G) Direct comparison of the turn decisions with real and virtual impact points. Same errors (C) to virtual and real impact point, same response latency (D), same kinematics (E, duration of bending phase). (F, G) As with real objects, accurate starts to virtual landing points are also possible across wide ranges of distance (F) and turning angle (G). See Figs. S1, S2 for additional information, Movies S1, S2 for examples showing the turn decision, and how they were evaluated. See text for detailed statistics. ***p<0.001, n.s. not significant.

Next, we needed to critically examine if the decisions made to ballistic VIPs (Movie S2), as mimicked and rewarded by the system, were equivalent to those made with real falling objects (see Movie S1). Given the limitations in the temporal resolution of the screen it is striking that no aspect of the decisions was statistically different under real and virtual conditions (Fig. 3C-G, n=101 (real) and n=85 (virtual) starts). The accuracy of the aim achieved at the end of the turns (measured as introduced in Fig. 2) was not statistically different (p=0.506, Mann-Whitney; Fig. 3C) and always minimized the error to the impact point, regardless of whether it was real or virtual (difference from zero: p=0.903 (real), p=0.455 (virtual); One-Sample Signed Rank) with also no significant difference in variance (p=0.589, Brown-Forsythe). The turns also had the same latency (between onset of prey motion and onset of the turns; p=0.297; Mann-Whitney; Fig. 3D) and kinematics (Fig. 3E: no difference in bending duration under real and VR conditions, p=0.071, Mann-Whitney). Furthermore, responses came from a large range of orientations and distances and, most importantly, were equally accurate across all distances (Fig. 3F) and turn sizes (Fig. 3G). In summary, despite the limited frame rate of the screens and despite the need to use highly conspicuous feeders in the majority of the (rewarded) presentations, the turn decisions were equivalent in all examined aspects (accuracy, latency, kinematics) to those obtained with real falling objects. In other words, the setup was sufficient to start our endeavor and allowed to study the archerfish’s highspeed decision exclusively with virtual stimuli.

Challenging the decision with a new rule that links input and rewarded output

The VR setup fulfilled all conditions required: it robustly elicits turn decisions that deviate in no detectable way from those elicited with real falling objects and that enable the fish to turn to the later ballistic landing point, real or virtual, from a wide range of distances and for all required turn sizes. This allowed us, to use the VR setup to explore the central question of the present study: would the decision be capable of adjusting to a novel rule of how the input variables allow to infer the rewarded virtual impact points? If the fish continued using ballistics to infer the VIP, as predicted, then their starts would now always be in error. The fish could then still learn to efficiently correct their initial errors later during their actual approach trajectory. Figure 4A introduces the new rule that we chose to face the fish with. It generates a systematic and substantial deviation (of 14.9 cm) between the new VIPs and the ballistic VIPs as inferred from the initial movement. After having travelled straight throughout what was the previous decision time, the movement direction is deflected by 39.8° (to the left, s. Methods) and the reward is now presented at the appropriate ‘deflected’ VIP. According to this new rule the connection between initial movement and the point of reward is still precisely defined. It would, therefore, in principle be possible to extract and use this new rule to infer the point of reward from the initial pre-deflection movement. Specifically, it would be easy to critically test whether the fish (or at least some of them) were using the new rule, by testing them exclusively with the initial movement (i.e., without the later deflection), and to measure if the turns are directed towards the calculated deflected or towards the calculated ballistic VIPs. Because the model behind our new rule might appear simple it is important to stress that learning to turn according to this new rule is not at all simple. It would require the fish to learn non-trivial corrections that can be appreciated by considering Fig. 4B: Relative to the turn they had previously made for a given input constellation they would need to add a correction that is not constant but depends sensitively on the position of the responding fish. This map of corrections is, furthermore, not the same for all input constellations. Instead, a new map of the kind shown in Fig. 4B is required for every combination of input variables (as illustrated with a simple example in Fig. S3).

The highspeed decision is not hardwired to ballistics but can be trained to a new rule.

(A) Sketch of the movement pattern used to introduce new virtual impact points (VIPs) that are still strictly defined by initial motion but systematically differ from the ballistic VIPs. Trajectories were initially straight but – after the time it typically takes the fish to make their decision – then changed their direction as illustrated. Once this new movement pattern was introduced, reward was given exclusively at the corresponding new ‘deflected’ VIPs and no longer at the ballistic VIP. (B) Learning to adjust the turns to initial prey motion requires non-trivial corrections to the previous turning angles (to the ballistic VIPs). These corrections depend on the fish’s initial position and a different map is needed for each set of motion input variables (as illustrated in Fig. S3). (C, D) Median errors of the turns relative to the new (C) and to the old ballistic (D) VIPs in successive training blocks 1 to 9 (each with 60 evaluated turn decisions), with food always delivered at the new VIPs. Colored background highlights when turns were oriented towards the new (C) and when to the old ballistic (D) VIPs. (E, F) Cumulative density functions (CDFs) at the blocks indicated by colored circles in (C, D) to show how the systematic error to the new VIPs systematically decreased (E) whereas errors to the ballistic landing points increased (F). (G) Errors made in several critical tests – from left to right: the deflection was visible and feeders were present at the new VIPs, just as during training (n=240), interspersed tests with only the initial straight trajectory shown (i.e. prey vanished before the deflection would occur) (n=68), interspersed tests (rate: 1 in 8) with also only the initial straight trajectory shown but direction of motion offset from the direction of the feeders by either 15° (n=197) or by 72° (n=145) to test if fish turned to the new VIP or to the feeder positions. (H, I) After training the fish were able to again respond appropriately over a wide range of distances (H) and required angles of turn (I). (J, K) Absence of speed-accuracy tradeoff of the re-programmed highspeed decisions. Plot of magnitude of error in aim versus response latency in the critical tests shown in (G) with only initial pre-deflection prey trajectory and offset of either 15° (J) or 72° (K). Note absence of significant correlation between error magnitude and response latency (p>0.3; R2<0.005). Data are represented as median in (C, D, G). ***p<0.001, **p<0.01, n.s. not significant.

If adult archerfish used a hardwired decision circuit, they could learn to rapidly modify their subsequent approach path, or they could abandon their rapid decisions altogether and instead use slower responses. To examine this, we now exclusively challenged the fish for many weeks with the new deflected trajectories and rewarded them at the newly defined VIPs. During this time, we continuously sampled the occurrence and error of all turn decisions (Figs. 4C, D). Interestingly, the fish never ceased to respond with their characteristic turns (response probability remained constant across all training blocks, p>0.08, Pairwise Chi-square). The kinematics of the turns and latency also remained constant (bending and straightening phase of their C-starts both p>0.28, latency p>0.07; Kruskal-Wallis). Furthermore, their aims were systematically in error for the initial training blocks (deviation from zero error to the deflected VIPs p≤0.001, One-Sample Signed Rank) as expected (Fig. 4D; deviation of aim to ballistic VIPs from zero error p>0.06; One-Sample Signed Rank; Movie S3). However, gradually the errors made to the new ‘deflected’ VIPs became smaller and, after more than 1750 presentations, the turns clearly minimized the error to the appropriate new ‘deflected’ VIPs (Figs. 4C, E, F; Movie S4) and no longer to the ballistic ones.

This surprising finding does by no means imply that the decision has now been based on the new rule. For example, the fish could have learned to feed information about the deflection of the prey’s trajectory into their unfolding turns and so achieve their new final turning angles, even though they had initiated their turns before the deflections were visible. To test this possibility, we interspersed tests in which the trained fish were only shown the initial linear part (i.e., the moving prey disappeared prior to the deflection) among many presentations with the full trajectory (including the deflection). The errors made in the interspersed tests (n=68) were not different from those made when the full trajectories were visible (Fig. 4G). This shows that in these critical tests the starts were still only made using the initial motion and did not require visibility of the deflection. Another way to explain the finding, however, could be that now the fish have learned to take the feeder positions into account, even though they did not in the previous experiments (Fig. 3B). In this scenario, the fish could have learned to use initial movement as an indicator which feeder to approach. To examine this possibility, we interspersed tests in which the direction of initial target movement was always either 15° or 72° off from the feeder position and in which, additionally, exclusively the initial movement was shown (i.e., prey vanished before the deflection). Furthermore, feeders were always offset both from the ballistic as well as from the new VIPs. Yet, the turns the fish made in these tests were still aimed at the new deflected VIPs (differences from zero error p=0.721 at 15° (n=197) and p=0.113 at 72° (n=145); One-Sample t-tests) but not at the feeders and also not at the ballistic VIPs (Kruskal-Wallis: p≤0.001 at 15°; p=0.005 at 72°; Dunn’s p<0.05 for comparison of errors; Movie S5). Furthermore, the turns – made when only the linear initial motion was visible and when the VIPs were spatially offset from the feeders – did not differ in latency (p=0.686, Kruskal-Wallis) nor in kinematics (bending duration: p=0.242, Kruskal-Wallis) from the turns made without offset and with the full trajectory visible (n=240).

Finally, the new turns were not restricted to just a few positions and orientations of the responding fish. Rather, they were possible and equally accurate across a range of distances and of turn angles (Fig. 4H, I). We also demonstrate another important feature, the absence of a speed-accuracy tradeoff also in the re-programmed decisions. The absence has previously been described for real, ballistically falling prey (11). In Fig. 4J, K we plot the magnitude of the error in the aim (immediately at the end of the C-starts, see Fig. 2B and Fig. S1) against response latency for the critical tests of Fig. 4G, in which only the linear initial part of motion was shown and offset by either 15° or 72° from the direction of the feeders. For both plots there was no significant linear (p=0.318, R2=0.005, Fig. 4J; p=0.469, R2=0.004, Fig. 4K) or Spearman correlation (p=0.128 and 0.174, respectively, in Fig. 4J, Fig. 4K). Our findings therefore demonstrate that a highspeed decision that lacks a speed-accuracy tradeoff can be flexibly reprogrammed to a new rule.

Immediate generalization to untrained input constellations

We next wondered about the degree of abstraction, if any, that was involved in reprogramming the decision. This can be examined conveniently, because all training to the new rule was carried out exclusively with targets presented at only one height level. In addition, possible target speed levels were limited, and prey was of only one size. In principle, the decision network could be retrained by simply replacing the angle required in the previous turns, given the position of the fish and the initial motion of prey (speed, direction, height) with the new angle. Formally this could be described as simply overwriting a ‘lookup table’. In this scenario the fish would not detect the underlying rule in a way that would enable it to generalize appropriately to novel untrained settings. So, in initial tests with motion shown at untrained height levels, predictably wrong turns should occur, because a lookup table would then either only contain (i) the former entries that were appropriate for ballistic motion (which we will call the prediction ‘ballistic’), or (ii) all its former ballistic entries could be substituted with the entries that were appropriate at the training height (which we will call the prediction ‘substitute’). Both predictions can be tested quantitatively, by calculating the turns the fish would have to make when they first encounter targets at untrained height levels (see Figs. 5A for an illustration of the predictions). For the actual tests, the untrained height levels were chosen such that the predicted errors to any of the correct VIPs would be at least 6 cm (and thus readily detectable). The actual errors made relative to the appropriate actual VIP are shown for the first 30 turns made with the novel larger (Fig. 5B) or novel lower (Fig. 5C) height level. Interestingly, the turns already minimized the error to the VIPs that were appropriate for the untrained novel height levels. They were appropriate right from start with no trend that errors would decrease (linear regression: all p>0.1). A detailed analysis of the errors (Figs. 5D, E) showed that the turns minimized the error to the deflected landing point also at the untrained height levels (difference from zero: both p>0.07; One-Sample t-test). They were also not directed to the ballistic landing point (prediction ‘ballistic’; both p<0.001; One-Sample Signed Rank) and not towards the deflected landing point at the training height (prediction ‘substitute’; both p<0.001; One-Sample Signed Rank). Again, none of the characteristics of the turns had changed in the novel situation compared to those after training (also n=30 right before these tests): latency: p=0.098, kinematics: p=0.177 (Kruskal-Wallis), variance of errors: p=0.188 (Brown-Forsythe). In other words, the new rule is not represented by updating a simple lookup table but in a way that allows immediate generalization to the new untrained height levels.

The new rule is represented in a way that allows immediate generalization.

(A) Idea for generalization tests that are possible because training to the ‘deflected’ trajectories (Fig. 4) was only for one level of target height. When faced with a larger than training height, fish would only be able to turn to the corresponding impact point P if they had acquired a more general rule of how to connect input (height, direction, speed) to the rewarding turn. Had they only exchanged those input-output experiences that were no longer rewarded, then they should still use the previously rewarded input-output relations and turn to the ballistic impact point (red, ‘’ballistic’’; see text). Had they substituted the old input-output relation for all height levels with the trained new ones, then they should turn to point ‘’substitute’’ (blue, see text) that would be appropriate for the training height. (B, C) When the fish were shown the deflected trajectories at larger (B) or lower (C) height than experienced during training, then already their first turns minimized the errors to the VIPs that are appropriate for the new rule at the untrained new height level with no indication that errors would initially be large and then decrease. (D, E) Closer analysis, using cumulative density functions (CDFs), for the errors made to the predicted points based on hypotheses introduced in (A), both for the larger (D) and for the lower (E) height (both n=30). (F) Evidence that learning the new rule had not used prior assumptions on target size. Errors of the turns made in the first 30 tests in which absolute target size was more than three-fold (13 mm) than the target size (4 mm) encountered throughout the training to the new rule. (G) CDFs of the first 30 tests with a low target speed (1.425 m/s) that the fish never encountered during training to the new rule. Again, the fish immediately chose turns to minimize the error to the predicted VIP based on the new rule but not the error to the points calculated based on predictions ‘’ballistic’’ or ‘’substitute’’.

We next tested whether re-programming of the decisions also allows generalization to untrained prey sizes and speed levels. Pursuing the same strategy as illustrated for height (Fig. 5A), we faced the trained fish with either a new absolute prey size or with new levels of prey speed. Again, the decision circuit could then use the previous entries in the hypothetical lookup table for these sizes and speed levels, that it would have from previous encounters with real, ballistically falling objects. Alternatively, it could have substituted the ballistic values with those that were appropriate during the training in Fig. 4. Again, the trained fish immediately turned to the deflected landing points (differences from zero error: each p>0.4, One-Sample Signed Rank; difference in aim before and after size change p=0.764, Mann-Whitney) and the error distributions were not significantly different (difference in distributions in Fig. 5F: p=0.278, Brown-Forsythe). Hence, learning did also not involve simplifying a priori assumptions about absolute prey size. The same held for untrained speed levels. When tests with lower than training speed were interspersed among presentations with the speed levels as during training, then already the first 30 turns made at the lower speed (Fig. 5G) minimized the error to the actual deflected VIP (difference from zero p=0.102; One-Sample t-test), but not to the VIPs given by either ballistics (prediction ‘ballistic’; p<0.001; One-Sample Signed Rank) or by using predictions for the deflected VIPs that are based on the trained speed levels (p<0.001; One-Sample Signed Rank). Comparing latency, bending duration and the variance of errors (speed: last 30 presentations with only training speed; size: last 30 presentations with target size during training) showed that changing target size and speed also did not affect the nature of the turns. Latency and bending duration were not significantly different in the changed versus training conditions (at novel untrained speed: Latency: p=0.280, Mann-Whitney; Stage 1: p=0.076, Mann-Whitney. At novel untrained size: Latency: p=0.305, t-test; Stage 1: p=0.997, Mann-Whitney). Variance was even slightly reduced (and not increased) at the untrained target speed (p=0.021, Brown-Forsythe) and remained unchanged after the increase in target size (p=0.742, Brown-Forsythe). In summary, all tests show that the decisions had been reprogrammed in much more clever ways than predicted. Rather than independently changing the size of an unrewarding turn for each individual constellation – as would seem required because the input variables themselves are varied independently from each other – learning allowed immediate generalization.

The decision can conditionally operate on two different rules

The fish were now no longer using ballistics to infer the rewarded virtual impact point from initial prey motion. But would they immediately switch back to using ballistics as soon as they encountered objects whose reward point was defined by ballistics, as throughout the fish’s previous life? To explore this possibility, we randomly showed our trained fish two differently shaped objects: Either (i) the disk to which the fish responded with turns toward the rewarded deflected VIPs, or (ii) a ‘ballistic’ object (triangle) that moved straight and for which a reward was given at the ballistic VIP (Fig. 6A). In the experiments, one of the two objects was chosen randomly to appear in stationary position for 2 s, and then it started moving. Our prediction was that the occurrence of objects with ballistic VIPs would reset everything and that the fish would now only turn towards ballistic VIPs – both for triangles and for circles. However, in experiments to test this prediction, the turns made in response to the ‘ballistic object’ (rewarded at the ballistic VIP) initially still minimized the error to the deflected VIP (difference from zero error p=0.256; One-Sample Signed Rank; same variance as for disk p=0.343; Brown-Forsythe; first 160 responses to ballistic object) and not to the ballistic VIP point (difference to ballistic landing point: p=0.038, Mann-Whitney; Fig. 6B,C). However, after 500 additional starts in response to the ballistic objects, turns were equally in error to the deflected and the ballistic VIPs (p=0.648, Mann-Whitney). In contrast, the turns made to the moving disks remained unaffected throughout the same period and remained perfectly matched to the deflected landing point (difference from zero error p=0.273; One-Sample Signed Rank; Fig. 6C). After about 300 additional encounters of moving triangles the turns minimized the error to the ballistic impact point when a triangle was shown (error to ballistic VIPs p=0.836, error to deflected VIPs p=0.020; One-Sample Signed Rank; Fig. 6B). However, the turns continued to minimize the error to the deflected impact point whenever the disk was shown (error to ballistic p≤0.001, error to deflected p=0.084, n=187; One-Sample Signed Rank; Fig. 6C).

The highspeed decision can conditionally use two different rules of how to connect input and output.

(A) After successful training to infer the new (deflected) virtual impact points (VIPs), fish were randomly shown either the disks with the deflected trajectories (reward at deflected VIPs) or new non-disk objects that moved on straight trajectories and were rewarded at the corresponding ballistic VIPs. (B, C) Error in the turns as determined in interspersed tests that only showed the straight initial movement both for non-disk objects (triangle, B) and for the familiar disk objects (C). Objects were chosen randomly. ‘Pre’ (filled symbols) denotes baseline before fish were exposed to a total of 500 presentations of the non-disk objects with reward at their ballistic VIPs. Errors were determined in subsequent test phases. As the fish changed their turn decision to minimize the error to the non-disk VIPs (B) they continued to aim to the deflected VIPs whenever they encountered moving disks (C) showing that they had not reversed to generally using ballistics again. (D-F) Interspersed tests that only showed the same initial linear trajectory but with either a disk (n=60) or a triangle moving (n=187). Graphs on right side to illustrate choice situation as seen from above: For the same set of input variables (speed, height, direction) the fish must turn to the ballistic VIP when a triangle is moving, but to the deflected VIP when a circle is moving. Detailed analysis shows that turns were appropriate to the correct VIP with no difference in error (D), latency (E) or kinematics (bending duration) (F). (G) Individual fish were able to select the appropriate turn to the deflected VIPs when encountering moving disks and to the ballistic VIP when encountering triangles. Aims were equally appropriate and did not differ in variability. Data are represented as medians (B, C, G). n.s. not significant.

Next, we interspersed tests that only showed the initial straight paths both for the disk (n=60) and for the triangle (n=187), i.e., the objects vanished before the deflection would occur for the disks. Picking either the ‘ballistic’ or the ‘deflected’ rule to choose the appropriate turn could now only be based on target shape. Again, the turns were toward the ballistic landing point when the triangle was shown (difference from ballistic p=0.075, from deflected p<0.001; One-Sample Signed Rank) and to the deflected landing point when the disk was shown (difference from ballistic p=0.008, from deflected p=0.412; One-Sample Signed Rank) (Fig. 6D). Also, latency (p=0.189, Mann-Whitney; Fig. 6E) and kinematics of the turns (bending duration; p=0.984; Mann-Whitney; Fig. F) were the same in the responses to triangles and disks. In other words, the nature of the turn decision was the same, but object shape determined which rule the decision used to determine the appropriate turn given the input data.

In principle these results could have arisen rather simply: some fish could have only responded to disks and others only to triangles, so that no single given individual would have to be able to handle both rules. To check if at least one individual would indeed be able to simultaneously use both rules we analyzed the contributions of the individual fish. Our analysis showed that all three of the three individual fish that had contributed most of the starts in Fig. 6 minimized the error to the appropriate VIP (Fig. 6G; difference from zero error always p>0.1; One-Sample Signed Rank). Also, the variability of the errors in applying the two rules did also not differ in any of the individuals (p=0.966; Brown-Forsythe). This shows clearly that the decision circuit used by archerfish does have the potential of simultaneously operating on two different rules.

A shape cue is needed to select which rule to use but need not be available a priori

In the previous series of experiments the fish could always see the shape of their virtual prey before it started to move. In principle, shape information could therefore have acted as a potential prior that would inform the decision circuit which of the two rules to later use when target movement starts. To test whether the circuit relied on such prior information we interspersed tests in which the prey object (disk or triangle) revealed its shape only at motion onset (Fig. 7). In each of these interspersed tests, first a neutral shape (a cross) appeared, remained stationary for 2s, and then was exchanged randomly with either the moving triangle or the moving disk. Moreover, only the straight initial path was shown so that shape after motion onset was the only cue available for selecting either the ‘ballistic’ or the ‘deflected’ rule (Movies S6, S7). Under these conditions, the decisions remained appropriate to the ballistic (triangle; n=109 starts) or deflected (disk; n=46 starts) VIPs (difference from zero error to appropriate VIP: p=0.451 (triangle) and p=0.671 (disk); p≤0.001 for the alternatives, all One-Sample Signed Rank). Furthermore, also the distribution of errors was not statistically different from those obtained with the shape cue available 2s prior to the decision (p=0.269; Brown-Forsythe). Hence the information that is needed to select which rule is to be used can also be extracted in the very brief interval after motion onset, together with all other decision-relevant information. It is not required to instruct the decision circuit before target movement starts.

The clue that signals which rule must be used is not required before prey motion starts.

In this series of experiments one of two differently shaped types of moving prey are shown, as in Fig. 6. Either a triangle that required the fish to turn to the appropriate ballistic VIPs, or a disk, that required the fish to turn not according to ballistics but based on the deflected rule (see Fig. 4). One of the two objects were randomly selected. In one type of experiments (’2s before’), the decision circuit was given time to adjust to the rule that was needed by displaying the object in stationary position for 2s prior to motion onset. In the other type (’only during decision phase’) no prior information was given. Instead, a neutral symbol (a cross) was shown 2s before prey motion and then the chosen object (disk or triangle) appeared and immediately was moving. In all tests, only the initial straight movement was shown (i.e., no deflection was visible in the moving disks). Errors were the same, regardless of whether object identity was revealed before or only during the very brief decision time (see Movies S6, S7). Data are represented as medians. n.s. not significant.

Discussion

Using the rapid turn decisions of hunting archerfish as an example we demonstrate that highspeed decisions that demonstrably lack a speed-accuracy tradeoff (Fig. 4J,K, 11), do not necessarily lack flexibility. Using a virtual reality system that exploits key properties of this decision (Figs. 2, 3) we were able to show that the decision is not hardwired to ballistic falling patterns. Before, the fish would turn to the later ballistic landing point, based on the initial values of prey motion. After a training phase in which the connection between initial motion and the point of reward had been systematically changed – according to a new rule – the fish were able to rapidly turn towards the new points, that were appropriate for their starting position and the combination of initial motion variables. Generalization tests after this reprogramming showed, furthermore, that the decision could immediately handle new constellations never encountered during the re-training phase and that it was capable of simultaneously handling two distinctly different rules, one for each of two visually different prey objects. Our findings provide a striking counterexample to the view that highspeed decision making is generally based on simplifying heuristics and lacks flexibility.

Comparing cognitive aspects of highspeed decision making with those of much slower behaviors

None of the cognitive aspects we describe here is in any way unique and not found in much slower behaviors. For example, archerfish generalize efficiently when they learn how to engage moving targets and to judge the absolute size of prey (44,45). Also, recognizing and assigning a value to different visual shapes has been demonstrated much earlier (e.g. 46) and many aspects have been described across the animal kingdom, including animals with far smaller brains (47–49). What is surprising, therefore, is clearly not to find cognitive features in archerfish, but rather to find such features in a decision that is made at a very different timescale at reflex-like speed (less than 100 ms) and that lacks a speed-accuracy tradeoff (Fig. 4J,K; 11). The complexity involved in reprogramming the archerfish highspeed decision must also not shy a comparison with tasks that are considered highly sophisticated in many laboratory animals (e.g., 50-52). For example, a much-studied task that is considered demanding is to have an animal under condition ‘a’ respond to a stimulus ‘α’ by turning left and right to stimulus ‘β’. Under condition ‘b’, however, the animal must do the opposite and turn right to stimulus ‘α’ and left to stimulus ‘β’. In comparison, the task the decision circuit of archerfish can handle in as little as just 40 milliseconds (see Fig. 4) is not only to conditionally switch between two motor outputs but to switch between two complete rules of how to connect a three-dimensional set of input data to a continuum of possible motor outputs.

Highspeed decision-making and minibrains

The plausible assumption that reflex-speed implies reduced complexity and flexibility (52,53) could have a similar foundation as the assertion that smaller brains have reduced cognitive capacities. Substantial progress has been made on this latter assertion, resulting in an impressive and continuously growing list of sophisticated cognitive features that were found in animals with bird- and fish brains (54–56). Most impressively, it has been demonstrated that insect brains that operate with a millionth of the neurons of a human brain can handle a wide range of cognitive tasks (47–49, 57–59) that even include social or observational learning (59–61). Of particular interest in our context is that it has even been demonstrated that learning can lead to the reprogramming of innate and supposedly hardwired behaviors in insects (62).

The reflex-speed of some decisions made by animals with much larger brains also will likely limit the size of the underlying neuronal circuitry, allowing only a subset of the brain to be used, a subset that could be considered a ‘functional minibrain’. The assertion that everything that is as fast as a reflex must be simple might therefore suffer the same fate as did the assertion that brains with fewer neurons must be simple. In the turn decisions of archerfish, specifically, it is known that the kinematics of their turn decisions is completely equivalent to the so-called C-starts that the fish produce to escape (40). Both the turn decisions (also called predictive C-starts; 34) and the archerfish escape C-starts are among the fastest C-start known in fish (40) and so at least the motor circuits are restricted to the well-known powerful hindbrain circuits of a comparably small number of tractable numbers of neurons, organized around the giant Mauthner neuron (63–66). Its rapidly conducting axon would be suited to initiate the turn but the recruitment of other hindbrain neurons is needed to account for the precision and adjusted straightening speed of the fish (40).

Are the findings typical for archerfish?

Our aim was to test the idea that the decisions are hardwired. To show that the decision has the potential to be reprogrammed – and that it can even operate on two rules – requires careful experiments and controls, but clearly one fish that can do it illustrates what the circuitry can do in principle. To demonstrate that the physiology of a tiger allows it to be green, observing one such tiger is sufficient. This is why, for example, studies on a single grey parrot (e.g., 67) have been extremely valuable. Here we analyzed a group of six fish that reached homogeneous stages of performances. Clearly, there could be large differences in the learning speed of the individuals. An extremely exciting possibility – that we cannot presently address – is if the reprogramming even would allow for some form of observational or even social learning. This would be highly valuable to speed up the learning and to reprogram the decisions with fewer examples.

We do believe, although the number of individuals that we were able to train and test in the present long endeavor is limited, that our findings are of relevance in the wild. First, the rapid and precise turn decisions are required because of the severe competition archerfish often face. Other surface-feeding fish are not only more numerous in Thailand – and therefore likely to be closer at the landing point of the archerfish’s prey – but they are better equipped with mechanosensors and are also capable of using vision to guide pre-impact turns. It is therefore crucial for archerfish to outperform them by using fast and precise turn decisions (33). We suggest that the clearly demonstrated capacity of these decisions for being flexible has evolved because of this competition and because of the scarcity of aerial prey at least in many natural habitats. This requires the fish to down and efficiently secure prey of various sizes. While small prey will typically fall ballistically (12,37), larger prey will face considerable frictional forces and – all other conditions being equal – will therefore not fall equally far. Simply using a hardwired circuitry for ballistically falling prey would produce wrong initial turns for these larger items and would require subsequent corrections that cost precious time and might mean losing some of the advantage over the competitors. The flexibility we found and the ability to operate on at least two types of prey that fall according to two different rules would allow at least some individual archerfish to efficiently handle two types of prey that co-occur during a specific time of the year. When one is replaced by another type of prey with yet another way in which its initial movement relates to its impact point, the decision would be able to adjust, again, to the new prey type, being still able to accurately respond to the prey that remained. So, we suggest, clearly without proof, that the ecological constraints and the severe competition that archerfish face in the wild could have favored the evolution of a decision that has the potential for plasticity and for combining speed, accuracy, and flexibility.

Conclusions

Our findings demonstrate that highspeed decisions that lack a speed-accuracy tradeoff are not necessarily limited to operating based on simplifying heuristics but can, at least in principle, offer learning capacities and cognitive aspects of much slower decisions and behaviors that occur at least at 100 times lesser speed. Our findings raise several exciting possibilities. First, it will be exciting to see if the capacities we found in archerfish are special or perhaps much more widespread among other highspeed decisions. Second, if highspeed decisions are not generally incapable of learning and flexibility, the obvious question remains more pressing what might be the other reasons that most decisions do show clear speed-accuracy tradeoffs (1). Third, the findings suggest that rapid decision-making can be efficiently used in many technological applications in which time is premium – even if such applications require a high degree of flexibility.

Materials and methods

Animals

Experiments were conducted on a group of six adult archerfish (Toxotes chatareus). The experimental fish were randomly chosen from a larger group of (male and female) archerfish. They had been imported from Thailand (Aquarium Glaser, Rodgau, Germany) and were kept at 26°C on a 12h:12h light/dark cycle in the large experimental tank (1m x 1m x 0.6 m, filled to a height of 34 cm) in brackish water (3.5 mS cm-1). All procedures were conducted in accordance with the German Animal Welfare Act (Tierschutzgesetz) as well as the German Regulation for the protection of animals used for experimental purposes or other scientific purposes (Tierschutz-Versuchstierordnung) and were approved by the government of Lower Franconia (Regierung Unterfranken, Würzburg, Germany).

General setup

To facilitate recording the fish from below – through the transparent bottom of the tank – together with the visual stimuli shown on the LCD monitor above, the tank was placed on steel frame stands, 110 cm above ground. A construction made from 3 cm aluminum profiles (item Industrietechnik GmbH, Germany) was used to hold the monitor on top of the tank as well as to position the feeders that were employed to reward the fish. To increase the contrast in the recordings of the fish, three conventional 500 W halogen floodlights, located around the tank, illuminated a square of white cloth (2 x 2 m) that was placed parallel to the water surface and 106 cm above it.

Monitors and stimuli

To present visual stimuli a 27’’ LCD monitor (Samsung SyncMaster S27A750D, 120 fps, max. intensity 300 cd/m2) was used. A glass plate in front of the screen protected it against shots of brackish water fired by the fish. All objects used in this study had a Michelson contrast of 0.96, as determined with a luminance meter (LS-110 Konica Minolta Sensing Europe) under the conditions of the actual experiments. The surface of the LCD monitor was parallel to the water surface, h = 28.8 cm above it and its center was directly above the tank’s center. Stimuli were generated using Apple Motion 5.4.7 and stored as 120 fps FullHD ProRes movies. We presented the movies with the presentation mode of Quicktime Player 10.4 (Apple Inc.) running on a MacPro 5.1 under MacOS 10.11. An AMD 4GB Radeon RX 480 graphic card produced the monitor signal and was connected via DisplayPort with the LCD monitor. Unless stated otherwise each presentation started with an object appearing in the center of the screen. After 2s the object starts moving in one of four pre-assigned, randomly chosen directions. Objects displayed were either disks of 4 mm diameter, disks of 13 mm diameter or equilateral triangles (height 13 mm). Objects that mimicked ballistically falling motion moved at a horizontal speed vh = 1.775 m s-1 and their virtual landing occurred 243 ms later, after a (virtual) path of 43 cm. On the screen, the movement of the ballistic objects was visible for 116 ms. Objects on the ‘deflected’ trajectories moved at the same horizontal speed but changed the direction of motion after being visible for 116 ms (or 20.5 cm) at a (virtual) time of 118.7 ms (i.e. after an initial straight path of dr = 21.1 cm) by an angle ɣ = 39.8° to the left (see Fig. 4A). They were then visible for a further 58 ms. The systematic change in direction causes the virtual impact point, the one inferred from initial motion, to be systematically offset by o = 2*(vh*sqrt(2h/g) - dr) * sin(ɣ/2) where h is the height at motion onset, g is the gravitational constant and dr is the distance travelled prior to the change in direction. In our experiments we selected conditions so that the offset was large enough (14.9 cm) to allow us detecting any errors in the bearing of the C-starts. In critical tests only the straight initial movement (i.e. the initial path of 20.5 cm) was shown both for ballistic and deflected trajectories.

Stimulus-coupled feeders

Unless stated otherwise 4 feeders were typically present. During training they were located at the places of possible virtual impact points. In many tests, the virtual impact points were systematically offset from the feeders by changing the direction of motion shown on the screen. This way the virtual landing points differed both from the position of the feeders as well as from the position relative to landmarks the fish might have been using during training. The feeders were custom-made (Fig S2A) and consisted of a tube (2 cm inner diameter, 44 cm long) with a sliding plate inserted 20 cm below its upper end to block the passage of a food pellet (one half of a Sera Cichlid stick). Each feeder was equipped with an electromagnet that can retract the slider to allow the passage of the pellet. All feeders were pre-loaded at random intervals before the tests, using a funnel (6 cm diameter) on the top side of each feeder (Fig S2A, B). Which of the four electromagnets was activated at which time was defined by trigger pules on the channels of the audio track of the movies shown on the LCD monitor so as to achieve an impact of the pellet at the virtual impact point and after the expected time T=sqrt(2h/g) of falling (h=initial height, g=gravitational acceleration). Four audio-channels were used to control which of the four electromagnets was to be activated. To achieve the required number of channels we connected an AV-receiver (Sony STR-DH520) to the computer via HDMI. Four output channels of the AV-receiver delivered the trigger signals to a custom-made amplifier which activated the electromagnets by means of a 200 ms pulse (12 V/300 mA). An additional audio channel triggered a pulse generator (TGP110; TTi, England) that started the highspeed video recording.

Highspeed imaging

All recordings were made from below through the tanks transparent bottom (Fig S2C, D). A Fastcam MC2 color 500 (Photron, Japan) highspeed camera, operated at 500 fps (512 x 512 pixels) and equipped with a CVM0411ND f1.6/4.4-11 mm lens (Lensation, Germany) was used to monitor the complete tank. The pre-trigger option of the camera was always used and set to record 4 s before and 4 s after the trigger signal (i.e. the onset of target motion). An Apple Macbook Pro 6.2 running Microsoft Windows 7 controlled the setup via Fastcam viewer 332. Recordings were stored as Quicktime movies and contained the appearance of the object on the screen, its motion, and the impact of food delivered by one of the feeders.

Discriminating the responses of individual group members

Because all experiments on the start decisions require the competition within a group (11) we confirmed that all fish of the group always contributed to the decisions reported throughout the study. This was checked at every stage of the work. Details on the individual performance are reported for the last and most demanding task that shows that these individuals had indeed learned everything that we report in this study (Fig. 6). Individuals were recognized based on the contour of their frontal body as seen from below. A template-contour was used for each individual fish that was drawn starting at one pectoral fin, over the snout and ending at the other pectoral fin. Respective images of the highspeed recordings were extracted using Image J (63) just when the straightening phase (so-called stage 2) of the so-called C-start (Fig. S1A) was terminated. Note that the turns are made right below the water surface and thus in a fixed distance from the camera. The contours taken at the end of stage 2 were specific for each individual fish in the experimental group. Discrimination performed this way coincided with one obtained independently by additionally recording the responses from the side (Photron Fastcam MC2 color 2K, equipped with same type of lens and also operated at 500 fps, view orthogonal to the front side and 50 cm away from it) in which the individuals could be identified by means of their individual stripe-patterns.

Natural stimuli used to test the virtual reality approach

A non-transparent platform (4.5 cm in diameter) was mounted above the center of the tank. This allowed to blow food pellets off the platform from a similar initial height as the virtual objects shown on the screen. Movement was started by an airflow (as in 11,37) and pellets fell ballistically toward the water surface.

Quantitative analysis of the turn responses

Image J 1.50 (68) was used to analyze the recorded videos. To ensure that the decisions could not be influenced by those of other fish, we exclusively analyzed the turn of the fish that responded first in each given test. Latency was the time between onset of movement of the prey object on the LED screen and the onset of the C-start. Error in aim relative to a point of interest (e.g., the virtual impact point) was determined from the last frame at the end of the straightening phase (stage 2) of the C-start, i.e., before the fish starts its approach trajectory (Fig. S1A). Orientation of the fish’s rigid front end was determined from the tip of the snout and the point midway between the two pectoral fins. If a line drawn along this orientation did cut a second line drawn between the start of motion and the point of interest (e.g., the virtual impact point), both projected to the water surface, then the error was defined negative, otherwise it was defined positive (Fig. S1B). Additionally, we determined kinematic aspects of the C-starts, such as the duration of its first bending stage (stage 1 of the C-start), the mean angular speed (i.e., turn angle / turn duration), and the initial orientation and distance the fish had before the onset of their C-start (40). Only C-starts that were completed before prey impact were included in our analysis, to ensure that the fish could not have used information from the impact of the food pellets to select their C-starts. Furthermore, only C-starts were included in which the subsequent approach path of the responding fish was not blocked by another fish. When aims had to be evaluated relative to two points of interest (e.g., virtual landing point and feeder position) then (i) both possible approach paths had to be free and not blocked by any group member and (ii) the responding fish had not to be oriented in a line with both points. Analysis of the coordinates and frame numbers determined with ImageJ was carried out using Excel (14.0 & 15.11).

Statistics

Statistics were run on GraphPadPrism 9.5. (GraphPad Software, USA). Tests referred to in the main text were as follows: Normality of data was assayed using Shapiro–Wilk tests. For single data sets (error of aim), difference from zero was checked using One-Sample t-tests (parametric) or One-Sample Signed Rank tests (non-parametric). Comparisons of two data sets (latency, duration of stage 1, error of aim) were performed with unpaired t-tests or Mann-Whitney tests. Comparisons of three or more data sets (latency, duration of stage 1, error of aim) were performed with Kruskal– Wallis One-Way ANOVA on ranks, post hoc tested with Dunn’s method (non-parametric). Equal variance between data sets (error of aim) was checked using Brown-Forsythe tests (non-parametric). Stability of aim was tested using linear regression and Spearman’s correlation. Tests for changes in C-start probability employed pairwise comparisons with χ2 tests. All respective tests were performed two-sided. ***p<0.001, **p<0.01, n.s. not significant. All data are available in the main text or the supplemental information.